---
layout: lesson
title: Introduction to mapping in R
subtitle: EDH7916
author: Benjamin Skinner
order: 6
category: extra
links:
  script: mapping.R
  pdf: mapping.pdf
  data: geo.zip
output:
  md_document:
    variant: gfm
    preserve_yaml: true
---

In this lesson, we'll go over the basics of mapping in R. As with most
lessons in this course, we'll only scratch the surface of what's
possible. Mainly, we will cover four things:

1. Understanding the basics of spatial data and using the `sf` library
   to work with it
1. Projections
1. Basic spatial joins
1. Making figures using `sf` and `ggplot2`

Importantly, we won't cover spatial statistics, which is a related,
but entirely other topic. At the end of this lesson, you should be
able to read in spatial data, perform some common data wrangling
procedures, and make a map that can be used to visualize spatial
relationships. 

Let's start!

```{r, include = FALSE, purl = FALSE}
source('knit_setup.R')
```
```{r, include = FALSE, purl = TRUE}
################################################################################
##
## <PROJ> EDH7916: Introduction to mapping in R
## <FILE> mapping.R 
## <INIT> 29 March 2022
## <AUTH> Benjamin Skinner (GitHub/Twitter: @btskinner)
##
################################################################################

```

# Libraries and paths

In addition to tidyverse, we'll use the `sf` library and `usmap`, the
latter of which provides some nicely formatted geospatial data for the
United States. Before you start the lesson, you'll need to install
both libraries: `install.packages(c("sf", "usmap"))`.

**NOTE** Once upon a time, getting the requisite spatial software
installed on your computer could be very difficult, with lots of separate
installations that could be quite finicky. But these days --- unless
you have a strange installation of R --- you should be able to just
install the `sf` library and have things work. If you run into
trouble, it's probably some background program that needs to be
updated or installed. If you end up with errors, copy and paste them
into Google. You will almost certainly find others who've had the same
problem and, hopefully (!), the solution. You can also check out the
[sf library webpage](https://r-spatial.github.io/sf/).

```{r}
## ---------------------------
## libraries
## ---------------------------

library(tidyverse)
library(sf)
library(usmap)
```

As usual, we'll set our directory paths. Notice that we have some
nested folders.

```{r}
## ---------------------------
## directory paths
## ---------------------------

## assume we're running this script from the ./scripts subdirectory
dat_dir <- file.path("..", "data")
geo_dir <- file.path(dat_dir, "geo")
usa_dir <- file.path(geo_dir, "cb_2018_us_state_20m")
```

# Projections and choropleths

In this first section, we'll work with a map of the United States to
look at differences in Bachelor's degree attainment across the
states. But before we read in the data, let's talk about a common
format for geographic data files, the shapefile.

First of all, what is a shapefile? While we can always rely on
Wikipedia for the [technical
answer](https://en.wikipedia.org/wiki/Shapefile), the gist is that
it's a special data structure that, in addition to the normal data
frames we are used to working with, contains geometric
information. This geographic information includes points, lines, and
polygons that can be used to create multidimensional figures such as
maps or to perform spatial analyses. Linking these geometric features
with the data frame, a researcher can describe and analyze the spatial
dimensions of data. For example, here are some questions that are
similar to some that I've explored in the past: 

- How close is this high school to the nearest open admissions college?
- How many colleges are within 10/50/100 miles of this census block?
- Are there differences in household earnings or average degree
  attainment between these proximate zip codes?

Of course, there are many more types of spatial questions to ask. If
you have a research question that has a spatial dimension, then you
are likely to use a shapefile or another spatial data format (we use
another type below). If you look inside the `cb_2018_us_state`
directory that's in `data/geo/`, you'll see a number of files with the
same name but different file endings:

```
./data/geo/cb_2018_us_state_20m/
├── cb_2018_us_state_20m.cpg
├── cb_2018_us_state_20m.dbf
├── cb_2018_us_state_20m.prj
├── cb_2018_us_state_20m.shp
├── cb_2018_us_state_20m.shp.ea.iso.xml
├── cb_2018_us_state_20m.shp.iso.xml
└── cb_2018_us_state_20m.shx
```

While the file ending in `*.shp` is technically the "shapefile," the
other files contain important information and are linked to the
primary shapefile. In other words, think of the collection of files as
the shapefile. So keep in mind: in the future when you download and/or
share a shapefile --- often in a `*.zip` format --- you need to
keep/include all the other files. If you are curious about what the
other files do, [check out this ArcGIS
page](https://desktop.arcgis.com/en/arcmap/latest/manage-data/shapefiles/shapefile-file-extensions.htm).

### Read in the data

Let's read in the data. Notice that we use a special function from the
sf library, `st_read()` and that we use the `*.shp` file ending. Other
than that, this should look like most of the other read functions
we've used before. Pipes from the tidyverse work here, so after
reading in the data, we'll go ahead and lower the data frame column
names (to save our pinkies from all the shift key work!) and convert
the [state FIPS
codes](https://en.wikipedia.org/wiki/Federal_Information_Processing_Standard_state_code),
which uniquely identify each state, from character strings to
integers. This will make our lives easier later when we filter our
data set.

```{r}
## ---------------------------
## input data
## ---------------------------

## reading in the least detailed cartographic boundary (cb) file from 2018
df_us <- st_read(file.path(usa_dir, "cb_2018_us_state_20m.shp")) %>%
  ## renaming the data frame columns to all lowercase
  rename_all(tolower) %>%
  ## converting state fips codes (two digit) to numbers for later filtering
  mutate(statefp = as.integer(statefp))
```

As it's read in, we get a little more information that tells us:

- The file path and driver (what's understanding the data structure):
`ESRI Shapefile`
- That it's a `Simple feature collection`, that is, a special data
  structure with `52` features (unique items: points, polygons, etc)
  and `9` fields (columns of information)
- The `Geometry type`, which is this case is a `MULTIPOLYGON`
- The coordinate dimension, which are `xy` here (think 2D)
- How big the full area covered is, in the scale of the underlying
  data. Since the default for this data file is geodesic (on the
  globe), we get longitude (x) and latitude (y) values
- The Coordinate Reference System (CRS), which determines how (if at
  all) the shapes are projected. In this case, they are not --- they
  remain specified as if on a globe --- with a datum (center point
  inside the globe) `NAD83`

We'll talk a little more about projections below. First, let's call
the data frame, `df_us`, by itself.

```{r}
## show
df_us
```

We get the same information as above, but now we can see the data
frame, too. This shapefile is pretty spare, giving us some basic
identifying data, such as the state name and land and water area
within the boundaries. In the last column, you can see the `geometry`,
which has the geometry type (`MULTIPOLYGON` here), as well the
beginnings of some numbers nested in parentheses. This is the extra
geometry information that we can use to make our maps.

### Mapping and changing projections

Let's make our first map. The sf library is powerful because it
smoothly integrates with ggplot. The new `geom_*` we'll use is
`geom_sf()`, which tells ggplot what it needs to know to make the
map. Because we're making a map and don't really care about axes or
all that (at least in this instance), we'll also add `theme_void()` at
the end of the call to remove everything but the map we plot.

```{r}
## ---------------------------
## initial plot
## ---------------------------

p <- ggplot(df_us) +
  geom_sf() +
  theme_void()

## show
p
```

Ok...

It's a map, but not super useful. What's going on? Well --- depending
on your screen resolution --- you might be able to tell that this map
of the United States includes various territories. Because these span
the globe, the map has to fit a huge spherical area into a smaller,
flatter space. This is what we get.

Since our plan is look at degree attainment in the states, we'll
reduce our data to only include states. We'll also start by removing
Alaska and Hawaii so that we can first practice with the lower 48 or
contiguous states.

```{r}
## ---------------------------
## lower 48 only plot
## ---------------------------

## filter to keep only states (fips <= 56, Wyoming) and then drop AK/HI
df_l48 <- df_us %>%
  filter(statefp <= 56) %>%
  filter(statefp != 02 & statefp != 15)

## same plot with reduced data
p <- ggplot(df_l48) +
  geom_sf() +
  theme_void()

## show
p
```

Now that we've limited our data and effectively zoomed in on the lower 48
states, our map is clearer and much more useful.

We have another issue, however. If you are used to looking at maps of
the United States, our figure looks a little flat and elongated. The
US - Canada border is very straight and the western part of the
country looks sort of stretched. This is because our data was
_unprojected_ and sf had to pick a projection in order to make the
figure.


What do we mean by _projection_? The fundamental problem of mapping is
this: we live on a spherical globe (technically closer to an [oblate
spheroid](https://en.wikipedia.org/wiki/Figure_of_the_Earth)) but want
to put maps on a flat, 2D surface. Imagine peeling the skin off an
orange in one piece and trying to lay it flat. If you want the peel to
have the same size and everything to roughly line up, then you'll have
[gaps between the
edges](https://en.wikipedia.org/wiki/Goode_homolosine_projection). Alternately,
you can remove the gaps and keep straight lines by stretching the
peel, but then you have a [stretched
surface](https://en.wikipedia.org/wiki/Mercator_projection)! In other
words, there's always a compromise. Most common projections try to
[preserve one
metric](https://en.wikipedia.org/wiki/Map_projection#Projections_by_preservation_of_a_metric_property):

- Conformal (preserve angles)
- Equal area (preserve area)
- Equidistant (preserve distances between points)

There is no single correct projection: it really depends on what you
want to do in your analyses. If you are calculating distances, then
you need to make sure your project preserves distance. If you want to
know about areas of overlapping shapes, then you will want to preserve
area with the projection you use. Check out [this
pdf](https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf)
for more information about projections or [this report from the
U.S. Geological Survey](https://pubs.usgs.gov/bul/1532/report.pdf) if
you _really_ want to know more!

Alternately, if you aren't doing any spatial analyses, then you can
just select the projection you find most aesthetically pleasing. When
faced with unprojected data, the sf library will by default choose an
[equirectangular
projection](https://en.wikipedia.org/wiki/Equirectangular_projection)
in which longitudinal and latitudinal lines are straight and at right
angles. This is how we get our stretched United States.

Let's choose a different Coordinate Reference System (CRS) so that our
map takes on a different shape. We'll use the USA Contiguous Albers
Equal Area Conic, which is good for North America and specifically the
lower 48 states. To reset the CRS, we'll use `st_transform()` with the
correct
[EPSG](https://en.wikipedia.org/wiki/EPSG_Geodetic_Parameter_Dataset)
numeric code for the projection we want: 5070.

```{r}
## ---------------------------
## transform CRS (re-project)
## ---------------------------

## change CRS to something that looks better for US
df_l48 <- st_transform(df_l48, crs = 5070)

## same plot, but should look different b/c we changed projection
p <- ggplot(df_l48) +
  geom_sf() +
  theme_void()

## show
p
```

Much nicer!

> #### Quick exercise 
> Re-project and re-plot your lower 48 states map `df_l48` using a CRS
> code of 3338 (good for Alaska) and 32136 (good for Tennessee). What
> happens to the map each time?

Of course, we dropped Alaska and Hawaii to make this figure and need
to bring them back. A common solution is to place them somewhere under
the southwestern states. We could do this ourselves, but it takes a
lot of manipulation of our the geometric data. The good news is that
someone has already done this work for us! Since we're only concerned
about state-level data, we'll use the `us_map()` function from the
**usmaps** library.

```{r}
## ---------------------------
## use usmaps to get AK/HI
## ---------------------------

## load usmap data for states
df_usmap <- us_map(regions = "states")
```

The usmaps library has its own plotting function that uses ggplot
under the hood, but since not all geospatial libraries have that,
we'll set up the data to work with sf. To do that, we'll use the
`st_as_sf()` function to set the underlying data as simple features
data and then combine the data points into polygons using a chain of
`group_by()`, `summarise()`, and then `st_cast()`. Run each line on
its own if you want to see what's happening with each step.

```{r}
## need to set up to work with sf
## start by setting up sf data frame with correct projection
df_usmap <- st_as_sf(df_usmap, coords = c("x", "y"), crs = 5070) %>%
  ## group by state / group (some states have non-contiguous parts like islands)
  group_by(fips, group) %>%
  ## combine these points into single geometry
  summarise(geometry = st_combine(geometry),
            .groups = "drop") %>%
  ## reset these points as polygons (think dots to lines that connect)
  st_cast("POLYGON")
```

Now that the data are set up correctly, we can run the same plotting
code as before, only changing the data. 

```{r}
## Alaska and Hawaii now included, but moved for tighter plot
p <- ggplot(df_usmap) +
  geom_sf() +
  theme_void()

## show
p
```

And now we have all the states in a nice projection, ready to make a
choropleth map.

### Choropleth

A choropleth map is just a map that uses color to show differences
across areas. Probably the most common choropleth maps people see are
during an election to show winners by political party in different
precincts or states. For our example, we'll show differences in
Bachelor's degree attainment rates for 25 to 44 year-olds from 2005 to
2019 across the states. First, we'll read in our data, which come from
the U.S. Census's American Community Survey.

```{r}
## ---------------------------
## BA attainment 2005-2019
## ---------------------------

## read in BA attainment (25-44yo) data
df_ba <- read_csv(file.path(geo_dir, "ba_25_44.csv"),
                  show_col_types = FALSE)

## show top of data
df_ba
```

First thing we need to do is join our new attainment data to our
spatial data frame. After creating a new join variable with the same
name and data type (`stfips = as.integer(fips)`), we can use
`left_join()` just as we've done in prior lessons.

```{r}
## join data to mapping data
df_usmap <- df_usmap %>%
  ## create a join variable, stfips, that matches what's in df_ba
  mutate(stfips = as.integer(fips)) %>%
  ## left_join as usual
  left_join(df_ba, by = "stfips")
```

For our first map, we'll filter to the latest year, 2019. So that the
color changes by BA percentage, we'll add `aes(fille = bpct)` in
`geom_sf()`. Everything else is the same as before.

```{r}
## plot with one year of BA attainment
p <- ggplot(df_usmap %>% filter(year == 2019)) +
  geom_sf(aes(fill = bapct)) +
  theme_void()

## show
p
```

To show changes over time, we'll filter in more years --- 2005, 2010,
2015, 2019 --- and include `facet_wrap(~ year)`, which will split our
plot into four panels, one for each year. Again, this is very similar
to plotting we've done before with ggplot.

```{r}
## plot with 4 years of BA attainment
p <- ggplot(df_usmap %>% filter(year %in% c(2005, 2010, 2015, 2019))) +
  facet_wrap(~ year) +
  geom_sf(aes(fill = bapct)) +
  theme_void()

## show
p
```

Just like our first figure, the differences are somewhat difficult to
see. Perhaps another color gradient would help. Alternately, we can
split our continuous percentages into a binary variable at some
threshold. Since a fairly common national statistic of Bachelor's
degree attainment is around 1/3 of adults, we'll pick 33% and use an
`ifelse()` statement to create `bacut`.

```{r}
## create new variable that's == 1 if BA attainment is >= 33%
df_usmap <- df_usmap %>%
  ## thinking ahead: use strings in yes/no options that will look good in legend
  mutate(bacut = ifelse(bapct >= 33, "33%+", "<33%"))
```

This time, we'll use our new variable `bacut` in the fill
aesthetic. We'll also make our legend nicer with a title using the
`name` option with `scale_fill_discrete()`.

```{r}
## plot with 4 years of BA attainment
p <- ggplot(df_usmap %>% filter(year %in% c(2005, 2010, 2015, 2019))) +
  facet_wrap(~year) +
  geom_sf(aes(fill = bacut)) +
  scale_fill_discrete(name = "BA attainment") +
  theme_void()

## show
p
```

This last figure makes cross-sectional and diachronic changes much
clearer. We could add further titles and a caption, but the figure
already looks pretty good!

> #### Quick exercise
> Choose a different cut point for BA attainment and different
> years. Plot another small multiples plots showing changes over
> time. 

# Spatial joins

In this second section, we'll perform a couple of simple spatial
joins. With spatial work, you'll often have one of these two
situations:

- Specific locations, like a school, that you want to place in its
  specific geographic location, such as its attendance zone or county
- Two different spatial polygons, like zip codes and census tracts,
  that don't perfectly align but that you want to intersect
  
We'll practice both of these procedures, which aren't too complicated
if your data are properly set up, meaning that you're using spatial
data sets that use the same projection (CRS). That's our case here, so
we're good to go. Instead of the entire country, we'll use school
attendance zones, school locations, and zip code areas in Alachua
County, Florida, to practice spatial joins.

First, we'll read it the data. Rather than using shapefiles, we'll use
[GeoJSON files](https://en.wikipedia.org/wiki/GeoJSON) this time,
which are a special type of JSON file (a plain text data file often
used in web applications) that contain geometric information.

## Read in data

```{r}
## ---------------------------
## Alachua County: school/zip
## ---------------------------

## school attendance zones; school locations; zip codes
df_zon <- st_read(file.path(geo_dir, "ac_school_zones.geojson"))
df_sch <- st_read(file.path(geo_dir, "ac_school_locs.geojson"))
df_zip <- st_read(file.path(geo_dir, "ac_zipcodes.geojson"))
```

## Points in polygons

First, we'll plot each set of school attendance zones with school
locations as points. We'll explicitly convert the school level
variable into a `factor()` so that we can order the plot from
elementary to middle to high school --- rather than alphabetical
order.

```{r}
## plot school zones + schools
p <- ggplot(df_zon) +
  ## convert school level to a factor so we can order it
  facet_wrap(~ factor(level, levels = c("elementary",
                                        "middle",
                                        "high"))) +
  ## change the color of zones by their code (just for variety)
  geom_sf(aes(fill = code)) +
  ## remove legend since it doesn't really tell us much
  scale_fill_discrete(guide = "none") +
  theme_void()

## show
p
```

Now that we've seen the attendance zones, let's plot the zip code
areas.

```{r}
## plot zip codes zones + schools
p <- ggplot(df_zip) +
  geom_sf(aes(fill = zip)) +
  scale_fill_discrete(guide = "none") +
  theme_void()

## show
p
```

Since zip codes are really about postal routes --- meaning they are
lines that have been converted into loose "areas" --- the overall
shape is a little strange and not as easily recognizable as Alachua
County.

Let's answer this question: _Which zip code contains the most
schools?_ To begin, we need to join our school data set to the zip
code data set. But instead of joining on a shared key like we've done
in prior lessons, we want schools to join based on the zip code they
_fall inside_.

To join school locations to the zip code areas in this way, we use
`st_join()` which is like `left_join()`, but takes into account the
`geometry` column in the data.

```{r}
## join schools to zip codes
df_sch_zip <- st_join(df_sch, df_zip)

## show
df_sch_zip
```

Now that we've placed schools in zip codes, we can answer the
question: _which zip code has the most schools_? Other than the second
line of code, `st_drop_geometry()`, which removes the geometry
features of the data set and makes it a plain data frame that's much
easier to manipulate, the rest of the dplyr chain should look
familiar. To answer our question, we:

1. `group_by()` zip column so that we are looking within zip code
1. `summarise()` and use `n()` to count the number of rows within each
   zip code, which, after our join, is the number of unique schools
1. `arrange()` using `desc()` so that the zip code with the maximum
   number is at the top of the printed data frame

```{r}
## what zip code has the most schools?
df_sch_zip %>%
  ## drop geometry so that we are left with simple data frame
  st_drop_geometry() %>%
  ## group by zip code
  group_by(zip) %>%
  ## use n() to get number of rows (i.e., schools) in each zip group
  summarise(num_schools = n()) %>%
  ## arrange results in descending order so that max is first
  arrange(desc(num_schools))
```

There's our answer!

> #### Quick exercise
> Rerun the last bit of code, but store the results in a tibble this
> time. Join the tibble back to the `df_zip` sf object and make a plot
> that color codes each zip code by the number of schools it contains.

## Overlapping polygons

For this last task, we'll zoom into a single school zone, Lake Forest
Elementary. We'll plot it first.

```{r}
## what zip code has the most schools?
p <- ggplot(df_zon %>% filter(facility == "Lake Forest Elementary")) +
  geom_sf() +
  theme_void()

## show
p
```

Knowing that zip codes and schools aren't perfectly aligned, that is,
a single school zone can be made up of multiple zip code areas and
vice versa, we ask: _how many zip codes are make up the Lake Forest
Elementary" attendance zone_?

Since we have two sets of polygons, we'll join using
`st_intersection()`. This particular join will use the left data frame
as the starting point and then keep the parts of the second data frame
that fit within (intersect) the first. Therefore, order matters here. 

Once we've done that, we'll plot the intersected data frame and fill
using `zip` so that we can see the zip code areas clearly.

```{r}
## join schools to zip codes
df_zon_zip <- st_intersection(df_zon, df_zip)

## what zip code has the most schools?
p <- ggplot(df_zon_zip %>% filter(facility == "Lake Forest Elementary")) +
  geom_sf(aes(fill = zip)) +
  scale_fill_discrete(name = "Zip Code") +
  theme_void()

## show
p
```

Now we can see that the Lake Forest Elementary attendance zone is made
up of three zip codes: mostly 32609 and 32641 with a little bit of
32640 in the southeast corner. We didn't project our data using an
area-preserving projection before doing these joins, so we would want
to be careful computing relative areas (though at this very small
area, the projected differences aren't likely to change much).

> #### Quick exercise
> Flip the order of `df_zon` and `df_zip` in `st_intersection()` in
> order to see how many school zones are in each zip code. Choose a
> zip code that covers more than one elementary zone and plot it,
> color coding the unique elementary school zones.

# Further information

As I said at the beginning of the lesson, we've only started with the
kinds of spatial analyses possible in R. If you want to learn more,
you might check out this [online resource](https://rspatial.org) or
[this book](http://gis.humboldt.edu/OLM/r/Spatial%20Analysis%20With%20R.pdf).

```{r, echo = FALSE, purl = TRUE}

## =============================================================================
## END SCRIPT
################################################################################
```
