---
layout: lesson
title: "Fitting regression models"
subtitle: "EDH7916 | Spring 2020"
author: Benjamin Skinner
order: 9
category: lesson
links:
  script: modeling.R
  pdf: modeling.pdf
  data: hsls_small.dta
output:
  md_document:
    variant: gfm
    preserve_yaml: true
---

```{r, include = FALSE, purl = FALSE}
source('knit_setup.R')
```
```{r, include = FALSE, purl = TRUE}
################################################################################
##
## <PROJ> EDH7916: Fitting regression models
## <FILE> modeling.R 
## <INIT> 24 March 2020
## <AUTH> Benjamin Skinner (GitHub/Twitter: @btskinner)
##
################################################################################
```

After your data have been wrangled from raw values to an analysis data
set and you've explored it with summary statistics and graphics, you
are ready to model it and begin making inferences. As one should
expect from a statistical language, R has a powerful system for
fitting statistical and econometric models.

In this lesson, we'll use data from the [NCES Education Longitudinal
Study of 2002](https://nces.ed.gov/surveys/els2002/). Much like HSLS,
ELS is a nationally representative survey that initially surveyed
students in their early high school career (10th grade) and followed
them into college and the workforce. We'll use a smaller version of it
because, unlike HSLS, the public use files for ELS include the
variables we need to properly account for the survey design when we
weight our data. Here's a codebook with descriptions of the variables
included in our lesson today:

| variable        |                                         description |
|:----------------|:----------------------------------------------------|
| stu\_id         |                                          student id |
| sch\_id         |                                           school id |
| strat\_id       |                                             stratum |
| psu             |                               primary sampling unit |
| bystuwt         |                                      student weight |
| bysex           |                                       sex-composite |
| byrace          |                  student's race/ethnicity-composite |
| bydob\_p        |                   student's year and month of birth |
| bypared         |                 parents' highest level of education |
| bymothed        |       mother's highest level of education-composite |
| byfathed        |       father's highest level of education-composite |
| byincome        | total family income from all sources 2001-composite |
| byses1          |                socio-economic status composite, v.1 |
| byses2          |                socio-economic status composite, v.2 |
| bystexp         | how far in school student thinks will get-composite |
| bynels2m        |    els-nels 1992 scale equated sophomore math score |
| bynels2r        | els-nels 1992 scale equated sophomore reading score |
| f1qwt           |                         questionnaire weight for f1 |
| f1pnlwt         |             panel weight, by and f1 (2002 and 2004) |
| f1psepln        |     f1 post-secondary plans right after high school |
| f2ps1sec        |           Sector of first postsecondary institution |
| female          |                                      == 1 if female |
| moth\_ba        |                            == 1 if mother has BA/BS |
| fath\_ba        |                            == 1 if father has BA/BS |
| par\_ba         |                     == 1 if either parent has BA/BS |
| plan\_col\_grad |        == 1 if student plans to earn college degree |
| lowinc          |                               == 1 if income < $25k |

Let's load the libraries and data!

...but first! You'll most likely need to install the [`survey`
package](https://CRAN.R-project.org/package=survey) first, 
using `install.packages("survey")`. 

```{r}
## ---------------------------
## libraries
## ---------------------------

library(tidyverse)
library(haven)
library(survey)
```

```{r}
## ---------------------------
## directory paths
## ---------------------------

## assume we're running this script from the ./scripts subdirectory
dat_dir <- file.path("..", "data")
```

```{r, purl = TRUE, include = FALSE}
## -----------------------------------------------------------------------------
## Model data
## -----------------------------------------------------------------------------
```

```{r}
## ---------------------------
## input data
## ---------------------------

## assume we're running this script from the ./scripts subdirectory
df <- read_dta(file.path(dat_dir, "els_plans.dta"))
```

# t-test
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## t-test
## ---------------------------
```

One common statistical test is a t-test for a difference in means
across groups (there are, of course, [others and R can compute
them](https://www.rdocumentation.org/packages/stats/versions/3.4.3/topics/t.test)). This
version of the test can be computed using the R formula syntax: `y
~ x`.  In our example, we'll compute base-year math scores against
mother's college education level. Notice that since we have the `data
= df` argument after the comma, we don't need to include `df$` before
the two variables.

```{r}
## t-test of difference in math scores across mother education (BA/BA or not)
t.test(bynels2m ~ moth_ba, data = df, var.equal = TRUE)
```

```{r, include = FALSE, purl = FALSE}
## t-test of difference in math scores across mother education (BA/BA or not)
ttest <- t.test(bynels2m ~ moth_ba, data = df, var.equal = TRUE)
```

> #### Quick exercise
> Run a t-test of reading scores against whether the father has a
> Bachelor's degree (`fath_ba`).

# Linear model
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## linear model
## ---------------------------

```
Linear models are the go-to method of making inferences for many data
analysts. In R, the `lm()` command is used to compute an OLS
regression. Unlike above, where we just let the `t.test()` output
print to the console, we can and will store the output in an object.

First, let's compute the same t-test but in a regression
framework. Because we assumed equal variances between the
distributions in the t-test above (`var.equal = TRUE`), we should get
the same results as we did before.

```{r}
## compute same test as above, but in a linear model
fit <- lm(bynels2m ~ moth_ba, data = df)
fit
```

The output is a little thin: just the coefficients. To see the full
range of information you want from regression output, use the
`summary()` function wrapped around the `fit` object.

```{r}
## use summary to see more information about regression
summary(fit)
```

Looks like the coefficient on `moth_ed`, ```r fit$coefficients[[2]]```,
is the same as the difference between the groups in the ttest, ```r 
ttest$estimate[[2]] - ttest$estimate[[1]]```, and the test statistics are
the same value: ```r ttest$statistic```. Success!

## Multiple regression
```{r, echo = FALSE, purl = TRUE}
## ------------
## lm w/terms
## ------------
```

To fit a multiple regression, use the same formula framework that
we've use before with the addition of all the terms you want on
right-hand side of the equation separated by plus (`+`) signs.

```{r}
## linear model with more than one covariate on the RHS
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc,
          data = df)
summary(fit)
```

The full output tells you:

* the model that you fit, under `Call:`
* a table of coefficients with
  * the main estimate (`Estimate`)
  * the estimate error (`Std. Error`)
  * the test statistic (`t value` with this model)
  * the p value (`Pr(>|t|`)
* significance stars (`.` and `*`) along with legend
* the R-squared values (`Multiple R-squared` and `Adjusted
    R-squared`)
* the model F-statistic (`F-statistic`)	
* number of observations dropped if any

If observations were dropped, you can recover the number of
observations used with the `nobs()` function.

```{r}
## check number of observations
nobs(fit)
```

The `fit` object also holds a lot of other information that is
sometimes useful.

```{r}
## see what fit object holds
names(fit)
```

For example, both `fitted.values` and `residuals` are stored in the
object. You can access these "hidden" attributes by treating the `fit`
object like a data frame and using the `$` notation.

```{r}
## see first few fitted values and residuals
head(fit$fitted.values)
head(fit$residuals)
```

> #### Quick exercise
> Add the fitted values to the residuals and store in an object
> (`x`). Compare these values to the math scores in the data frame.

As a final note, the model matrix used fit the regression can be
retrieved using `model.matrix()`. Since we have a lot of observations,
we'll just look at the first few rows.

```{r}
## see the design matrix
head(model.matrix(fit))
```

What this shows is that the fit object actually stores a copy of the
data used to run it. That's really convenient if you want to save the
object to disk (with the `save()` function) so you can review the
regression results later. But keep in mind that if you share that
file, you are sharing the part of the data used to estimate it.

## Using categorical variables or factors
```{r, echo = FALSE, purl = TRUE}
## ------------
## factors
## ------------
```

It's not necessary to pre-construct dummy variables if you want to use
a categorical variable in your model. Instead you can use the
categorical variable wrapped in the `factor()` function. This tells R
that the underlying variable shouldn't be treated as a continuous
value, but should be discrete groups. R will make the dummy variables
on the fly when fitting the model. We'll include the categorical
variable `byrace` in this model.

```{r}
## add factors
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc + factor(byrace),
          data = df)
summary(fit)
```

If you're using labeled data like we have been for the past couple of
modules, you can use the `as_factor()` function from
the
[haven library](https://www.rdocumentation.org/packages/haven/versions/1.1.0/topics/as_factor) in
place of the base `factor()` function. You'll still see the
`as_factor(<var>)` prefix on each coefficient, but now you'll have
labels instead of the underlying values, which should make parsing the
output a little easier.

```{r}
## same model, but use as_factor() instead of factor() to use labels
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc + as_factor(byrace),
          data = df)
summary(fit)
```

If you look at the model matrix, you can see how R created the dummy
variables from `byrace`.

```{r}
## see what R did under the hood to convert categorical to dummies
head(model.matrix(fit))
```

> #### Quick exercise
> Add the categorical variable `byincome` to the model above. Next use
> `model.matrix()` to check the RHS matrix.

## Interactions
```{r, echo = FALSE, purl = TRUE}
## ------------
## interactions
## ------------
```

Add interactions to a regression using an asterisks (`*`) between the
terms you want to interact. This will add both main terms and the
interaction(s) between the two to the model. Any interaction terms
will be labeled using the base name or factor name of each term
joined by a colon (`:`).


```{r}
## add interactions
fit <- lm(bynels2m ~ byses1 + factor(bypared)*lowinc, data = df)
summary(fit)
```

## Polynomials
```{r, echo = FALSE, purl = TRUE}
## ------------
## polynomials
## ------------
```

To add quadratic and other polynomial terms to the model, use the
`I()` function, which lets you raise the term to the power you want in
the regression using the caret (`^`) operator. In the model below, we
add a quadratic version of the reading score to the right-hand side.

```{r}
## add polynomials
fit <- lm(bynels2m ~ bynels2r + I(bynels2r^2), data = df)
summary(fit)
```

> #### Quick exercise
> Fit a linear model with both interactions and a polynomial term. Then
> look at the model matrix to see what R did under the hood.

# Generalized linear model
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## generalized linear model
## ---------------------------
```

To fit a model with binary outcomes, switch to the `glm()`
function. It is set up just like `lm()`, but it has an extra argument,
`family`. Set the argument to `binomial()` when your dependent
variable is binary. By default, the `link` function is a
[logit](https://en.wikipedia.org/wiki/Logit) link.

```{r}
## logit
fit <- glm(plan_col_grad ~ bynels2m + as_factor(bypared),
           data = df,
           family = binomial())
summary(fit)
```

If you want a [probit](https://en.wikipedia.org/wiki/Probit_model)
model, just change the link to `probit`. 

```{r}
## probit
fit <- glm(plan_col_grad ~ bynels2m + as_factor(bypared),
           data = df,
           family = binomial(link = "probit"))
summary(fit)
```

> #### Quick exercise
> Fit a logit or probit model to another binary outcome.

# Using survey weights
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## survey weights
## ---------------------------
```

So far we haven't used survey weights, but they are very important
when using survey data. To use survey weights, you'll need to use the
survey package (which we've already loaded above)


To use survey weights, you need to set the survey design using the
`svydesign()` function. You could do this in the `svyglm()` function
we'll use to actually estimate the equation, but it's easier and
clearer to do it first, store it in an object, and then use that
object in the `syvglm()`.

ELS has a complex sampling design that we won't get into, but the
appropriate columns from our data frame, `df`, are set to the proper
arguments in `svydesign()`:

* `ids` are the primary sampling units or `psu`s  
* `strata` are indicated by the `strat_id`s  
* `weight` is the base-year student weight or `bystuwt`
* `data` is our data frame object, `df`
* `nest = TRUE` because the `psu`s are nested in `strat_id`s

Finally, notice the `~` before each column name, which is necessary in
this function.

```{r}
## set svy design data
svy_df <- svydesign(ids = ~psu,
                    strata = ~strat_id,
                    weight = ~bystuwt,
                    data = df,
                    nest = TRUE)
```

Now that we've set the survey design, let's compare the unweighted
mean with one that accounts for the survey design;

```{r}
## compare unweighted and survey-weighted mean of math scores
mean(df$bynels2m, na.rm = TRUE)
svymean(~bynels2m, design = svy_df, na.rm = TRUE)
```

We can also use the survey package to properly weight our t-tests and
regression models --- again, using the object `svy_df` in
the `design` argument in place of our unset `df` data frame.

```{r}
## get svymeans by group
svyby(~bynels2m, by = ~moth_ba, design = svy_df, FUN = svymean, na.rm = TRUE)

## t-test using survey design / weights
svyttest(bynels2m ~ moth_ba, design = svy_df, var.equal = TRUE)
```

> #### QUICK EXERCISE
> Compare this to the output from `t.test()` above.

```{r}
## fit the svyglm regression and show output
svyfit <- svyglm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc,
                 design = svy_df)
summary(svyfit)
```

Note that the survey library has a ton of features and is worth diving
into if you regularly work with survey data.

## Predictions
```{r, echo = FALSE, purl = TRUE}
## ------------
## predictions
## ------------
```

Being able to generate predictions from new data can be a powerful
tool. Above, we were able to return the predicted values from the fit
object. We can also use the `predict()` function to return the
standard error of the prediction in addition to the predicted values
for new observations.

First, we'll get predicted values using the original data along with
their standard errors.

```{r}
## predict from first model
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc,
          data = df)

## old data
fit_pred <- predict(fit, se.fit = TRUE)

## show options
names(fit_pred)
head(fit_pred$fit)
head(fit_pred$se.fit)
```
### Predictions with new data

Ideally, we would have a new observations with which to make
predictions. Then we could test our modeling choices by seeing how
well they predicted the outcomes of the new observations.

With discrete outcomes (like binary 0/1 data), for example, we could
use our model and right-hand side variables from new observations to
predict whether the new observation should have a 0 or 1
outcome. Then we could compare those predictions to the actual
observed outcomes by making a 2 by
2 [confusion matrix](https://en.wikipedia.org/wiki/Confusion_matrix)
that counted the numbers of true positives and negatives (correct
predictions) and false positives and negatives (incorrect
predictions). 

With continuous outcomes, we could follow the same procedure as above,
but rather than using a confusion matrix, instead assess our model
performance by measuring the error between our predictions and the
observed outcomes. Depending on our problem and model, we might care
about minimizing the root mean square error, the mean absolute error,
or some other metric of the error.

### Predictions using training and testing data

In the absence of new data, we instead could have separated our data
into two data sets,
a
[training set and test set](https://en.wikipedia.org/wiki/Training,_test,_and_validation_sets). After
fitting our model to the training data, we could have tested it by
following either above procedure with the testing data (depending on
the outcome type). Setting a rule
for ourselves, we could evaluate how well we did, that is, how well
our training data model classified test data outcomes, and perhaps
decide to adjust our modeling assumptions.

## Margins

Using the `predict()` function alongside some other skills we have
practiced, we can also make predictions on the margin a la
Stata's
[`-margins-` suite of commands](https://www.stata.com/help.cgi?margins).

For example, after fitting our multiple regression, we might ask
ourselves, what is the marginal "effect" of having a low family income
on math scores, holding all other terms in our model constant?

To answer this question, we first need to make a "new" data frame with
a column each for the variables used in the model and rows that equal
the number of predictive margins that we want to create. In our
example, that means making a data frame with two rows and five
columns.

With `lowinc`, the variable that we want to make marginal predictions
for, we have two potential values: 0 and 1. This is the reason our
"new" data frame has two rows. If `lowinc` took on four values, for
example, then our "new" data frame would have four rows, one for each
potential value. But since we have two, `lowinc` in our "new" data frame
will equal `0` in one row and `1` in the other row.

All other columns in the "new" data frame should have consistent
values down their rows. Often, each column's repeated value is the
variable's average in the data. Though we could use the original data
frame (`df`) to generate these averages, the resulting values may
summarize different data from what was used to fit the model if there
were observations that `lm()` dropped due to missing values. That
happened with our model. We could try to use the original data frame
and account for dropped observations, but I think it's easier to use
the design matrix that's retrieved from `model.matrix()`.

The code below goes step-by-step to make the "new" data frame.

```{r}
## create new data that has two rows, with averages and one marginal change

## (1) save model matrix
mm <- model.matrix(fit)
head(mm)

## (2) drop intercept column of ones (predict() doesn't need them)
mm <- mm[,-1]
head(mm)

## (3) convert to data frame so we can use $ notation in next step
mm <- as.data.frame(mm)

## (4) new data frame of means where only lowinc changes
new_df <- data.frame(byses1 = mean(mm$byses1),
                     female = mean(mm$female),
                     moth_ba = mean(mm$moth_ba),
                     fath_ba = mean(mm$fath_ba),
                     lowinc = c(0,1))

## see new data
new_df
```
Notice how the new data frame has the same terms that were used in the
original model, but has only two rows. In the `lowinc` column, the
values switch from `0` to `1`. All the other rows are averages of the
data used to fit the model.

To generate the prediction, we use the same function call as before,
but use our `new_df` object with the `newdata` argument.

```{r}
## predict margins
predict(fit, newdata = new_df, se.fit = TRUE)
```

Our results show that compared to otherwise similar students, those
with a family income less than $25,000 a year are predicted to score
about two points lower on their math test. 

In this example, we held the other covariates at their means. We could
have chosen other values (*e.g.* `fath_ba == 1` or `female == 1`),
however, meaning that we could use the same procedure to produce
predictions for low-income status (or other model covariates) across a
range of margins.


```{r, echo = FALSE, purl = TRUE}
## =============================================================================
## END SCRIPT
################################################################################
```
