---
layout: lesson
title: "Inferential II: Regression and prediction"
subtitle: EDH7916
author: Benjamin Skinner
order: 1
category: extra
links:
  script: modeling.R
  pdf: modeling.pdf
  data: els_plans.dta
output:
  md_document:
    variant: gfm
    preserve_yaml: true
---

```{r, include = FALSE, purl = FALSE}
source('knit_setup.R')
```
```{r, include = FALSE, purl = TRUE}
################################################################################
##
## <PROJ> EDH7916: Inferential II: Regression and prediction
## <FILE> regression.R 
## <INIT> 24 March 2020
## <AUTH> Benjamin Skinner (GitHub/Twitter: @btskinner)
##
################################################################################
```

In this lesson, we'll move beyond t.tests to regression. As before,
this lesson is too short to stand in for a full course on regression
analyses. However, it should give you the basics of how to fit
regression and store key parameters. We'll also cover the basics of
prediction and the estimate of marginal "effects" (nothing causal here!).

# Data

In this lesson, we'll use the same data from the [NCES Education
Longitudinal Study of 2002](https://nces.ed.gov/surveys/els2002/). So
you don't have to go back to the prior lesson, here again is a
codebook with descriptions of the variables included in our lesson
today:

| variable        |                                         description |
|:----------------|:----------------------------------------------------|
| stu\_id         |                                          student id |
| sch\_id         |                                           school id |
| strat\_id       |                                             stratum |
| psu             |                               primary sampling unit |
| bystuwt         |                                      student weight |
| bysex           |                                       sex-composite |
| byrace          |                  student's race/ethnicity-composite |
| bydob\_p        |                   student's year and month of birth |
| bypared         |                 parents' highest level of education |
| bymothed        |       mother's highest level of education-composite |
| byfathed        |       father's highest level of education-composite |
| byincome        | total family income from all sources 2001-composite |
| byses1          |                socio-economic status composite, v.1 |
| byses2          |                socio-economic status composite, v.2 |
| bystexp         | how far in school student thinks will get-composite |
| bynels2m        |    els-nels 1992 scale equated sophomore math score |
| bynels2r        | els-nels 1992 scale equated sophomore reading score |
| f1qwt           |                         questionnaire weight for f1 |
| f1pnlwt         |             panel weight, by and f1 (2002 and 2004) |
| f1psepln        |     f1 post-secondary plans right after high school |
| f2ps1sec        |           Sector of first postsecondary institution |
| female          |                                      == 1 if female |
| moth\_ba        |                            == 1 if mother has BA/BS |
| fath\_ba        |                            == 1 if father has BA/BS |
| par\_ba         |                     == 1 if either parent has BA/BS |
| plan\_col\_grad |        == 1 if student plans to earn college degree |
| lowinc          |                               == 1 if income < $25k |

We'll load the same libraries and data!  

```{r}
## ---------------------------
## libraries
## ---------------------------

library(tidyverse)
library(haven)
library(survey)
```

```{r}
## ---------------------------
## directory paths
## ---------------------------

## assume we're running this script from the ./scripts subdirectory
dat_dir <- file.path("..", "data")
```

```{r, purl = TRUE, include = FALSE}
## -----------------------------------------------------------------------------
## Model data
## -----------------------------------------------------------------------------
```

```{r}
## ---------------------------
## input data
## ---------------------------

## assume we're running this script from the ./scripts subdirectory
df <- read_dta(file.path(dat_dir, "els_plans.dta"))
```
# Linear model
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## linear model
## ---------------------------

```
Linear models are the go-to method of making inferences for many data
analysts. In R, the `lm()` command is used to compute an [ordinary
least squares (OLS)](https://en.wikipedia.org/wiki/Ordinary_least_squares)
regression. Unlike above, where we just let the `t.test()` output
print to the console, we can and will store the output in an object.

First, let's compute the same t-test as in the prior inferential
lesson, but in a regression framework. This time, we'll assume equal
variances between the distributions in the t-test above (`var.equal =
TRUE`), so we should get the same results as we did before.

```{r, include = FALSE, purl = FALSE}
## t-test of difference in math scores across mother education (BA/BA or not)
ttest <- t.test(bynels2m ~ par_ba, data = df, var.equal = TRUE)
```

```{r}
## t-test of difference in math scores across parental education (BA/BA or not)
t.test(bynels2m ~ par_ba, data = df, var.equal = TRUE)
```

```{r}
## compute same test as above, but in a linear model
fit <- lm(bynels2m ~ par_ba, data = df)
fit
```

The output is a little thin: just the coefficients. To see the full
range of information you want from regression output, use the
`summary()` function wrapped around the `fit` object.

```{r}
## use summary to see more information about regression
summary(fit)
```
We'll more fully discuss this output in the next section. For now, let's compare
the key findings to those returned from the t test.

Because our right-hand side (RHS) variable is an indicator variable
that `== 0` for parents without a BA/BS or higher and `== 1` if either
parent has a BA/BS or higher, then the intercept reflects the math
test score for students when `par_ba == 0`. This matches the `mean in
group 0` value from the t test above.

In a regression framework, the coefficient on `par_ba` is the marginal
difference when `par_ba` increases by one unit. Since `pared` is the
only parameter on the RHS (besides the intercept) and only takes on
values 0 and 1, we can add its coefficient to the intercept to get the
math test score mean for students with parents with a BA/BS or higher.

```{r}
## add intercept and par_ba coefficient
fit$coefficients[["(Intercept)"]] + fit$coefficients[["par_ba"]]
```

Looks like this value matches what we saw before (within
rounding). Going the other way, the coefficient on `par_ba`, ```r fit$coefficients[[2]]```, 
is the same as the difference between the
groups in the t test. Finally, notice that the absolute value of the
test statistic for the t test and the `par_ba` coefficient are the
same value: ```r abs(round(ttest$statistic, 2))```. Success!

## Multiple regression
```{r, echo = FALSE, purl = TRUE}
## ------------
## lm w/terms
## ------------
```

To fit a multiple regression, use the same formula framework that
we've use before with the addition of all the terms you want on
right-hand side of the equation separated by plus (`+`) signs.

_**NB** From here on out, we'll spend less time interpreting the
regression results so that we can focus on the tools of running
regressions. That said, let me know if you have questions of
interpretation._

```{r}
## linear model with more than one covariate on the RHS
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc,
          data = df)
summary(fit)
```

The full output tells you:

* the model that you fit, under `Call:`
* a table of coefficients with
  * the point estimates (`Estimate`)
  * the point estimate errors (`Std. Error`)
  * the test statistic for each point estimate (`t value` with this model)
  * the p value for each point estimate (`Pr(>|t|)`)
* significance stars (`.` and `*`) along with legend
* the R-squared values (`Multiple R-squared` and `Adjusted
    R-squared`)
* the model F-statistic (`F-statistic`)	
* number of observations dropped if any

If observations were dropped due to missing values (`lm()` does this
automatically by default), you can recover the number of observations
actually used with the `nobs()` function.

```{r}
## check number of observations
nobs(fit)
```

The `fit` object also holds a lot of other information that is
sometimes useful.

```{r}
## see what fit object holds
names(fit)
```

In addition to the `coefficients`, which you pulled out of the first
model, both `fitted.values` and `residuals` are stored in the
object. You can access these "hidden" attributes by treating the `fit`
object like a data frame and using the `$` notation.

```{r}
## see first few fitted values and residuals
head(fit$fitted.values)
head(fit$residuals)
```

> #### Quick exercise
> Add the fitted values to the residuals and store in an object
> (`x`). Compare these values to the math scores in the data frame.

As a final note, the model matrix used fit the regression can be
retrieved using `model.matrix()`. Since we have a lot of observations,
we'll just look at the first few rows.

```{r}
## see the design matrix
head(model.matrix(fit))
```

What this shows is that the fit object actually stores a copy of the
data used to run it. That's really convenient if you want to save the
object to disk (with the `save()` function) so you can review the
regression results later. But keep in mind that if you share that
file, you are sharing the part of the data used to estimate
it. Because a lot of education data is restricted in some way --- via
memorandums of understanding (MOUs) or IRB --- be careful about
sharing the saved output object. Typically you'll only share the
results in a table or figure, but just be aware.

## Using categorical variables or factors
```{r, echo = FALSE, purl = TRUE}
## ------------
## factors
## ------------
```

It's not necessary to pre-construct dummy variables if you want to use
a categorical variable in your model. Instead you can use the
categorical variable wrapped in the `factor()` function. This tells R
that the underlying variable shouldn't be treated as a continuous
value, but should be discrete groups. R will make the dummy variables
on the fly when fitting the model. We'll include the categorical
variable `bystexp` in this model. 

```{r}
## check values of student expectations
df %>%
    count(bystexp)
```

Even though student expectations of eventual degree attainment are
roughly ordered, let's use them in our model as discrete groups. That
way we can leave in "Don't know" and don't have to worry about that
"attend college, 4-year degree incomplete" is somehow _higher_ than
"attend or complete 2-year college/school".

```{r}
## add factors
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba
          + lowinc + factor(bystexp),
          data = df)
summary(fit)
```

If you're using labeled data like we have been for the past couple of
modules, you can use the `as_factor()` function from
the
[haven library](https://haven.tidyverse.org/reference/as_factor.html) in
place of the base `factor()` function. You'll still see the
`as_factor(<var>)` prefix on each coefficient, but now you'll have
labels instead of the underlying values, which should make parsing the
output a little easier.

```{r}
## same model, but use as_factor() instead of factor() to use labels
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba
          + lowinc + as_factor(bystexp),
          data = df)
summary(fit)
```

If you look at the model matrix, you can see how R created the dummy
variables from `bystexp`: adding new columns of only 0/1s that
correspond to the `bystexp` value of each student.

```{r}
## see what R did under the hood to convert categorical to dummies
head(model.matrix(fit))
```

> #### Quick exercise
> Add the categorical variable `byincome` to the model above. Next use
> `model.matrix()` to check the RHS matrix.

## Interactions
```{r, echo = FALSE, purl = TRUE}
## ------------
## interactions
## ------------
```

Add interactions to a regression using an asterisks (`*`) between the
terms you want to interact. This will add both main terms and the
interaction(s) between the two to the model. Any interaction terms
will be labeled using the base name or factor name of each term
joined by a colon (`:`).


```{r}
## add interactions
fit <- lm(bynels2m ~ byses1 + factor(bypared)*lowinc, data = df)
summary(fit)
```

## Polynomials
```{r, echo = FALSE, purl = TRUE}
## ------------
## polynomials
## ------------
```

To add quadratic and other polynomial terms to the model, use the
`I()` function, which lets you raise the term to the power you want in
the regression using the caret (`^`) operator. In the model below, we
add a quadratic version of the reading score to the right-hand side.

```{r}
## add polynomials
fit <- lm(bynels2m ~ bynels2r + I(bynels2r^2), data = df)
summary(fit)
```

> #### Quick exercise
> Fit a linear model with both interactions and a polynomial term. Then
> look at the model matrix to see what R did under the hood.

# Generalized linear model for binary outcomes
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## generalized linear model
## ---------------------------
```

In some cases when you have binary outcomes --- 0/1 --- it may be
appropriate to continue using regular OLS, fitting what is typically
called a [_linear probability model_ or
LPM](https://en.wikipedia.org/wiki/Linear_probability_model). In those
cases, just use `lm()` as you have been.

But in other cases, you'll want to fit a [generalized linear
model](https://en.wikipedia.org/wiki/Generalized_linear_model), in
which case you'll need to switch to the `glm()` function. It is set up
just like `lm()`, but it has an extra argument, `family`. Set the
argument to `binomial()` when your dependent variable is binary. By
default, the `link` function is a
[logit](https://en.wikipedia.org/wiki/Logit) link.

```{r}
## logit
fit <- glm(plan_col_grad ~ bynels2m + as_factor(bypared),
           data = df,
           family = binomial())
summary(fit)
```

If you want a [probit](https://en.wikipedia.org/wiki/Probit_model)
model, just change the link to `probit`. 

```{r}
## probit
fit <- glm(plan_col_grad ~ bynels2m + as_factor(bypared),
           data = df,
           family = binomial(link = "probit"))
summary(fit)
```

Note that the interpretation of parameters in logit and probit models
differs from that in linear model with a normal / gaussian link
function. For example, a one-unit change for a parameter in a logistic
regression is associated with a change in _log odds_ of the outcome, 
$$ 
log(\frac{p}{1-p}) 
$$ 
, holding all other values at some fixed value
(their means, for example). If this doesn't sound particularly
interpretable, it's not! This is one reason people continue to use
linear probability models (LPMs) rather than logits/probits (there are
[other
reasons](https://opa.hhs.gov/sites/default/files/2020-07/lpm-tabrief.pdf)
too). As always, make the choice that best fits your data, research
questions, and intended audience.

> #### Quick exercise
> Fit a logit or probit model to another binary outcome.


# Using survey weights in a regression
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## survey weights
## ---------------------------
```

We spent time in the Inferential I lesson on setting up a data
frame that accounted for survey weights. Review that if need a
reminder of the intuition. 

Important for this lesson is that you can (and should!) use survey
weights in a regression framework. As a reminder, here's the
information you need to set up the `svydesign()` that accounts for
[ELS's complex sampling design](https://nces.ed.gov/training/datauser/ELS_04.html):

* `ids` are the primary sampling units or `psu`s  
* `strata` are indicated by the `strat_id`s  
* `weight` is the base-year student weight or `bystuwt`
* `data` is our data frame object, `df`
* `nest = TRUE` because the `psu`s are nested in `strat_id`s

Before running our survey-weighted regression, we'll once again set up
our data in a new object.

```{r}
## subset data
svy_df <- df %>%
    select(psu,                         # primary sampling unit
           strat_id,                    # stratum ID
           bystuwt,                     # weight we want to use
           bynels2m,                    # variables we want...
           moth_ba,
           fath_ba,
           par_ba,
           byses1,
           lowinc,
           female) %>%
    ## go ahead and drop observations with missing values
    drop_na()

## set svy design data
svy_df <- svydesign(ids = ~psu,
                    strata = ~strat_id,
                    weight = ~bystuwt,
                    data = svy_df,
                    nest = TRUE)
```

Now that we've done that, here's how we run a weighted
regression. Notice that it's `svyglm()`, even though we used `lm()`
before (the default `"link"` function in `glm()` is `gaussian` or
normal; if we had binary outcomes and wanted to use a logit link then
we could include `family = binomial()` as we did before).

```{r}
## fit the svyglm regression and show output
svyfit <- svyglm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc,
                 design = svy_df)
summary(svyfit)
```

As when compared t.test results when using unweighted and weighted
data, our regression results are little different when using the
weights. As a reminder, there's no single quick answer to using
weights in a regression framework. It's up to you and your
investigation of the code book to decide:

- the appropriate weights to choose (there are usually many options!)
- what weights mean in the context of missing data
- what weights mean in a complex research design

## Predictions
```{r, echo = FALSE, purl = TRUE}
## ------------
## predictions
## ------------
```

Being able to generate predictions from new data can be a powerful
tool. Above, we were able to return the predicted values from the fit
object. We can also use the `predict()` function to return the
standard error of the prediction in addition to the predicted values
for new observations.

First, we'll get predicted values using the original data along with
their standard errors.

```{r}
## predict from first model
fit <- lm(bynels2m ~ byses1 + female + moth_ba + fath_ba + lowinc,
          data = df)

## old data
fit_pred <- predict(fit, se.fit = TRUE)

## show options
names(fit_pred)
head(fit_pred$fit)
head(fit_pred$se.fit)
```

With the standard errors, we can get a better feel for how much faith
we want to put in our predictions. If the standard errors are low,
then maybe we feel more secure than if the errors are
large. Alternatively, perhaps the errors are uniformly lower for some
parts of our data than others. If so, that might suggest more
investigation or a note on the limitations of our predictions for
parts of the sample.

### Two other types of predictions

We won't practice these since in application they work similarly to
what we did above. However, I do want to note two other types of
predictions that you might want to make. Both involve making
predictions for data that you didn't use to fit your
model. Predictions using new data or held-out data in a train/test
framework are another way to evaluate your model. If you predict well
to new/held-out data, that can be a good sign for the utility of your
model.

#### Predictions with new data

Ideally, we would have a new observations with which to make
predictions. Then we could test our modeling choices by seeing how
well they predicted the outcomes of the new observations.

With discrete outcomes (like binary 0/1 data), for example, we could
use our model and right-hand side variables from new observations to
predict whether the new observation should have a 0 or 1
outcome. Then we could compare those predictions to the actual
observed outcomes by making a 2 by
2 [confusion matrix](https://en.wikipedia.org/wiki/Confusion_matrix)
that counted the numbers of true positives and negatives (correct
predictions) and false positives and negatives (incorrect
predictions). 

With continuous outcomes, we could follow the same procedure as above,
but rather than using a confusion matrix, instead assess our model
performance by measuring the error between our predictions and the
observed outcomes. Depending on our problem and model, we might care
about minimizing the root mean square error, the mean absolute error,
or some other metric of the error.

#### Predictions using training and testing data

In the absence of new data, we instead could have separated our data
into two data sets, a [training set and test
set](https://en.wikipedia.org/wiki/Training,_test,_and_validation_sets). After
fitting our model to the training data, we could have tested it by
following either above procedure with the testing data (depending on
the outcome type). Setting a rule for ourselves, we could evaluate how
well we did, that is, how well our training data model classified test
data outcomes, and perhaps decide to adjust our modeling
assumptions. This is a fundamental way that many machine learning
algorithm assess fit.

## Margins

Using the `predict()` function alongside some other skills we have
practiced, we can also make predictions on the margin a la
Stata's
[`-margins-` suite of commands](https://www.stata.com/help.cgi?margins).

For example, after fitting our multiple regression, we might ask
ourselves, what is the marginal association of coming from a family
with low income on math scores, holding all other terms in our model
constant? In other words, if student A and student B are similar along
dimensions we can observe (let's say the average student in our
sample) except for the fact that student A's family is considered
lower income and student B's is not, what if any test score difference
might we expect?

To answer this question, we first need to make a "new" data frame with
a column each for the variables used in the model and rows that equal
the number of predictive margins that we want to create. In our
example, that means making a data frame with two rows and five
columns.

With `lowinc`, the variable that we want to make marginal predictions
for, we have two potential values: 0 and 1. This is the reason our
"new" data frame has two rows. If `lowinc` took on four values, for
example, then our "new" data frame would have four rows, one for each
potential value. But since we have two, `lowinc` in our "new" data frame
will equal `0` in one row and `1` in the other row.

All other columns in the "new" data frame should have consistent
values down their rows. Often, each column's repeated value is the
variable's average in the data. Though we could use the original data
frame (`df`) to generate these averages, the resulting values may
summarize different data from what was used to fit the model if there
were observations that `lm()` dropped due to missing values. That
happened with our model. We could try to use the original data frame
and account for dropped observations, but I think it's easier to use
the design matrix that's retrieved from `model.matrix()`.

The code below goes step-by-step to make the "new" data frame.

```{r}
## create new data that has two rows, with averages and one marginal change

## (1) save model matrix
mm <- model.matrix(fit)
head(mm)

## (2) drop intercept column of ones (predict() doesn't need them)
mm <- mm[,-1]
head(mm)

## (3) convert to data frame so we can use $ notation in next step
mm <- as_tibble(mm)

## (4) new data frame of means where only lowinc changes
new_df <- tibble(byses1 = mean(mm$byses1),
                 female = mean(mm$female),
                 moth_ba = mean(mm$moth_ba),
                 fath_ba = mean(mm$fath_ba),
                 lowinc = c(0,1))

## see new data
new_df
```
Notice how the new data frame has the same terms that were used in the
original model, but has only two rows. In the `lowinc` column, the
values switch from `0` to `1`. All the other rows are averages of the
data used to fit the model. This of course makes for a somewhat
nonsensical average (what does it mean that a single father to have
.32 of a BA/BS?), but that's okay. Again, what we want right now are
two students who represent the "average" student except one is low
income and the other is not.

To generate the prediction, we use the same function call as before,
but use our `new_df` object with the `newdata` argument.

```{r}
## predict margins
predict(fit, newdata = new_df, se.fit = TRUE)
```

Our results show that compared to otherwise similar students, those
with a family income less than $25,000 a year are predicted to score
about two points lower on their math test. To be clear, this is not a
casual estimate, but rather associational. This means we would be
wrong to say that having a lower family income _causes_ lower math
test scores (which doesn't jibe with our domain knowledge either; low
income is almost certainly a proxy for other omitted variables ---
access to educational resources for just one thing --- that are much
more directly linked to test scores).

I also want to note that we held the other covariates at their
means. We could have instead chosen other values (*e.g.* `fath_ba ==
1` or `female == 1`), which would have given us different marginal
associations. Dropping or including other covariates likely would
change our results, as well. The takeaway is that is that margins you
compute are a function of your model as well as (obviously) the margin
you investigate.

```{r, echo = FALSE, purl = TRUE}
## =============================================================================
## END SCRIPT
################################################################################
```
