---
layout: lesson
title: Web scraping in R
subtitle: EDH7916
author: Benjamin Skinner
order: 2
category: extra
links:
  script: scraping.R
  pdf: scraping.pdf
output:
  md_document:
    variant: gfm
    preserve_yaml: true
---

```{r, include = FALSE, purl = FALSE}
source('knit_setup.R')
```
```{r, include = FALSE, purl = TRUE}
################################################################################
##
## <PROJ> EDH7916: Web scraping
## <FILE> scraping.R 
## <INIT> 18 July 2020
## <AUTH> Benjamin Skinner (GitHub/Twitter: @btskinner)
##
################################################################################
```

This lesson introduces the basic steps to scrape data from a website
using the [rvest](https://CRAN.R-project.org/package=rvest). Because
there are about as many ways to scrape a website as there are types of
web data that you want to gather, web scraping is both art and
science, with varying degrees of data cleaning required. If you are
lucky, data will be regularly and unambiguously formatted, meaning
that it is easy to grab the data you want in the format that you
want. If you are less lucky, regular expressions to clean strings will
quickly become your friend.

Knowing a bit about web design, specifically
[HTML](https://www.w3schools.com/html/default.asp),
[XML](https://www.w3schools.com/xml/default.asp), and
[CSS](https://www.w3schools.com/css/default.asp) is helpful when web
scraping. This lesson focuses on static sites, but sites that require
user interaction (*e.g.*, clicking a button or inputting data into a
form in order to show data) can also be scraped. These sites require
special packages such as
[RSelenium](https://docs.ropensci.org/RSelenium/) and some knowledge
of [Javascript](https://www.w3schools.com/js/default.asp) is helpful.

For this lesson, however, we'll read static web tables from [NCES
Digest of Education
Statistics](https://nces.ed.gov/programs/digest/2017menu_tables.asp). NCES
helpfully makes these tables available in downloadable Excel
worksheets, but we'll pretend they don't exist for the
moment. Specifically, we'll focus on [Table
302.10](https://nces.ed.gov/programs/digest/d17/tables/dt17_302.10.asp),
which shows numbers of high school graduates and percentage of college
enrollment, broken out by gender and college level, for the years 1960
through 2016.

```{r}
## ---------------------------
## libraries
## ---------------------------

library(tidyverse)
library(rvest)
library(lubridate)
```

# Inspect the web site

First, let's check out the [table we want to
scrape](https://nces.ed.gov/programs/digest/d17/tables/dt17_302.10.asp). The
table we see looks like a regularly formatted table, much like we
would see in a paper document. But unlike a printed document, a web
page relies on hidden-from-the-user code to generate what we see. By
doing it this way instead of serving a static image, websites can
adjust to the wide array of user screen sizes, devices, and operating
systems. Instructions that tell the user device how to generate the
page are also smaller than sending a preformatted image, so bandwidth
and time to load are also reduced.

But as web scrapers, we don't need this. We need the underlying
HTML/CSS/XML code used to generate the page. To see it, you'll need to
use a web site inspector. With
[Firefox](https://developer.mozilla.org/en-US/docs/Tools/Page_Inspector)
and
[Chrome](https://developers.google.com/web/tools/chrome-devtools/),
you should be able to right-click the page and see the underlying code
(you may need to turn on developer tools first). With Safari, you will
have to [enable the developer tools
first](https://developers.google.com/web/tools/chrome-devtools/).

The top code of the page should look something like this:  

```html
<!DOCTYPE html PUBLIC "-//W3C//DTD XHTML 1.0 Transitional//EN" "http://www.w3.org/TR/xhtml1/DTD/xhtml1-transitional.dtd">
<!-- Current year pub navigation function -->

```

Moving further down, we find the table data, but in a very different
format (first row):  

```html
...
<tr>
  <th class="TblCls009" scope="row" nowrap="nowrap">1960 </th>
  <td class="TblCls010">1,679</td>
  <td class="TblCls011">(44.5)</td>
  <td class="TblCls010">756</td>
  <td class="TblCls011">(32.3)</td>
  <td class="TblCls010">923</td>
  <td class="TblCls011">(30.1)</td>
  <td class="TblCls010">45.1</td>
  <td class="TblCls011">(2.16)</td>
  <td class="TblCls010">&mdash;</td>
  <td class="TblCls011">(&dagger;)</td>
  <td class="TblCls010">&mdash;</td>
  <td class="TblCls011">(&dagger;)</td>
  <td class="TblCls010">54.0</td>
  <td class="TblCls011">(3.23)</td>
  <td class="TblCls010">&mdash;</td>
  <td class="TblCls011">(&dagger;)</td>
  <td class="TblCls010">&mdash;</td>
  <td class="TblCls011">(&dagger;)</td>
  <td class="TblCls010">37.9</td>
  <td class="TblCls011">(2.85)</td>
  <td class="TblCls010">&mdash;</td>
  <td class="TblCls011">(&dagger;)</td>
  <td class="TblCls010">&mdash;</td>
  <td class="TblCls011">(&dagger;)</td>
</tr>
...
```
The task is to convert these data into a data frame that we can then
store or use in tables and figures. This is what the
[rvest](https://CRAN.R-project.org/package=rvest) helps us do.

# Read web site
```{r, echo = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Read underlying web page code
## ---------------------------------------------------------
```

The first step is to read the web page code into an object using the
`read_html()` function.

```{r}
## set site
url <- "https://nces.ed.gov/programs/digest/d17/tables/dt17_302.10.asp"

## get site
site <- read_html(url)
```

Showing our object, we can see that the basic structure of the web
page is stored.

```{r}
## show
site
```

# Select nodes

Right now, we have a structured, but not particularly useful object
holding our web page data. To pull out specific data, we use the
`html_nodes()` function. Selecting a node is somewhat akin to using
dplyr's `filter()` on a data frame.

Great...but what's a node and how do I know which ones to use? First,
a node is a particular element that is comprised of some information
stored between, for example, HTML tags like `<p>...</p>` or
`<h1>...</h2>`. Good web design says that information on page should
be organized by its purpose and similarity to other data. For example,
major headers should be wrapped in `<h1>` tags and similar page
sections should be given the same CSS class. We can use CSS ids and
classes with the `html_nodes()` function to pull the exact data we
need.

Great!...but what are the classes that we need? Well, we could just
inspect the web page manually and guess. For some pages, that works
great. But it certainly looks like a chore for this page. Luckily,
there's a great tool that will help us.

## SelectorGadget

[SelectorGadget](http://selectorgadget.com) is a (now very old) plugin
that allows you to click on a web page and, through process of
elimination, get the exact combination of HTML tags and CSS ids and
classes you need to pull only the data you need.

You can get it from the link above or by [following the instructions
here](https://rvest.tidyverse.org/articles/selectorgadget.html). 

The SelectorGadget page has instructions, but briefly, this is the
process:  

1. On the first click, SelectorGadget will make its best guess about
   what you want based on the item you clicked (*e.g.*, table
   column). The particular element you clicked will be green. The
   other elements it assumed you want will turn yellow. Sometimes it's
   right and you're finished!  
2. Often, it will select something you don't want. In that case, click
   on the yellow item you don't want. Again, SelectorGadget will make
   and informed guess. Sometimes it will drop all extraneous elements
   and sometimes you will need to click multiple times. These elements
   will be red.
3. On the other hand, SelectorGadget may not have given you everything
   you want. Keep clicking on new elements (and dropping the extra)
   until only what you want is highlighted in either green or yellow.
   
As you're clicking, you'll see a box with a string of element ids and
classes changing. When you're finished, copy this string. This is your
node you'll use in the `html_nodes()` function!

> #### Quick exercise 
> Get the SelectorGadget plugin and play with it
> for a few minutes. See if you can select only a specific column then
> only a specific row.

# First column of data 

```{r, echo = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Get data for recent graduate totals
## ---------------------------------------------------------
```
As a first step, let's get the first column of data in Table 302.10:
the total number of recent high school graduates. Using
SelectorGadget, I see that the node string I should use is
`'.tableBracketRow td:nth-child(2)'`. After selecting the node, we use
`html_text()` to convert the data into a vector like we're used to
seeing.

```{r}
## subset to just first column
tot <- site %>%
  html_nodes(".tableBracketRow td:nth-child(2)") %>%
  html_text()

## show
tot
```

So far so good, but we can see a few problems. First, the blank rows
in the table show up in our data. While those blank table spaces are good
for the eyes, they aren't good in our data set. Let's try to remove
them using the `trim = TRUE` option.

```{r}
## ...this time trim blank spaces
tot <- site %>%
  html_nodes(".tableBracketRow td:nth-child(2)") %>%
  html_text(trim = TRUE)

## show
tot
```

Better, but the empty elements are still there. We can use
`str_subset()` from the stringr library (loaded with tidyverse) to
remove them.

```{r}
## remove blank values; str_subset removes pattern ("")
tot <- tot %>% str_subset(pattern = "")

## show
tot
```

Getting closer. Next, let's convert our numbers to actual numbers,
which R thinks are strings at the moment. To do this, we need to get
rid of the commas. The `str_replace()` function is perfect for this.
Regular expressions can become complicated, but our use
here is simple:

```{r}
## remove commas, replacing with empty string
tot <- tot %>% str_replace(pattern = ",", replacement = "")

## show
tot
```

Now we're ready to convert to a number.

```{r}
## convert to numeric
tot <- tot %>% as.integer()

## show
tot
```

Finished!

# Add year
```{r, echo = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Add years
## ---------------------------------------------------------
```

So that these numbers make sense, let's grab the years column and
create and data frame so that we can make a figure of long term high
school completer totals. Again, the first step is to use
SelectorGadget to get the node string. This time, it's `"tbody th"`.


```{r}
## get years column
years <- site %>%
  html_nodes("tbody th") %>%
  html_text(trim = TRUE)

## remove blank spaces like before
years <- years %>% str_subset(pattern = "")

## show
years
```

We've gotten rid of the blank items, but now we have a new problem:
the footnotes in the last few years has just be added to the
year. Instead of 2010, we have 20103, and so on through 2016. Since
the problem is small (it's easy to see all the bad items) and regular
(always extra 3 as the 5th digit), we can fix it using `str_sub()`.  

```{r}
## trim footnote that's become extra digit
years <- years %>% str_sub(start = 1, stop = 4)

## show
years
```

Fixed! Now we bind together with our high school completers
total. Because we want to make a time period line graph, we'll also
convert the years to a date format. We'll use `ymd` from the lubridate
library. Since we only have years, we'll include the argument
`truncated = 2L`, which means that we have an incomplete date (no
month or day).

**NB** Since we dropped blank elements in each vector separately, it's
important to check that all the data line up properly now that we've
bound them together. If we wanted to be safer, we could have bound the
data first, then dropped the rows with double missing values.

```{r}
## put in data frame
df <- bind_cols(years = years, total = tot) %>%
  mutate(years = ymd(years, truncated = 2L))

## show
df
```

You can see that the date format adds a month and day (January 1st by
default). While these particular dates probably aren't right, we won't
use them later when graphing so they can stay.

Let's plot our trends.

```{r scrape_trends_tot}
## plot
g <- ggplot(df, mapping = aes(x = years, y = total)) +
  ## line for the main estimate
  geom_line() +
  ## make x-axis look nice
  ## major breaks: every 5 years, from min year to max year
  ## minor breaks: every 1 year, from min year to max year
  ## labels: formate to only show year ("%Y")
  scale_x_date(breaks = seq(min(df$years),
                            max(df$years),
                            "5 years"),
               minor_breaks = seq(min(df$years),
                                  max(df$years),
                                  "1 years"),
               date_labels = "%Y") +
  ## nice labels and titles
  labs(x = "Year",
       y = "High school completers (1000s)",
       title = "Total number of high school completers: 1960 to 2016",
       caption = "Source: NCES Digest of Education Statistics, 2017, Table 302.10")
g
```

> #### Quick exercise 
> Pull in total percentage of enrollment (column 5), add to data
> frame, and plot against year.

# Scrape entire table
```{r, echo = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Scrape entire table
## ---------------------------------------------------------

```

Now that we've pulled two columns, let's try to grab the entire
table. Once again, we'll use SelectorGadget to get our node string.

```{r}
## save node
node <- paste0(".TblCls002 , td.TblCls005 , tbody .TblCls008 , ",
               ".TblCls009 , .TblCls011 , .TblCls010")

## save more dataframe-friendly column names that we
## get from looking at the table online
nms <- c("year","hs_comp_tot", "hs_comp_tot_se",
         "hs_comp_m", "hs_comp_m_se",
         "hs_comp_f", "hs_comp_f_se",
         "enr_pct", "enr_pct_se",
         "enr_pct_2", "enr_pct_2_se",
         "enr_pct_4", "enr_pct_4_se",
         "enr_pct_m", "enr_pct_m_se",
         "enr_pct_2_m", "enr_pct_2_m_se",
         "enr_pct_4_m", "enr_pct_4_m_se",
         "enr_pct_f", "enr_pct_f_se",
         "enr_pct_2_f", "enr_pct_2_f_se",
         "enr_pct_4_f", "enr_pct_4_f_se")

## whole table
tab <- site %>%
  ## use nodes
  html_nodes(node) %>%
  ## to text with trim
  html_text(trim = TRUE)

## show first few elements
tab[1:30]
```

Okay. It looks like we have it, but it's all in single dimension
vector. Since we eventually want a data frame, let's convert to a matrix.

```{r}
## convert to matrix
tab <- tab %>%
  ## we know the size by looking at the table online
  matrix(., ncol = 25, byrow = TRUE)

## dimensions
dim(tab)

## show first few columns using base R [<rows>,<cols>] notation
tab[1:10,1:5]
```

> #### Quick exercise
> What happens if you don't use `byrow = TRUE` in the matrix command?

It's getting better, but now we have a lot of special characters that
we need to clean out. This section relies more heavily on regular
expressions, but the idea is the same as above. 

```{r}
## clean up table
tab <- tab %>%
  ## convert to tibble, leaving name repair as minimal for now
  as_tibble(.name_repair = "minimal") %>%
  ## rename using names above
  set_names(nms) %>%
  ## remove commas
  mutate(across(everything(), ~ str_replace(., ",", ""))) %>%
  ## remove dagger and parentheses
  mutate(across(everything(), ~ str_replace_na(., "\\(\U2020\\)"))) %>%
  ## remove hyphens
  mutate(across(everything(), ~ str_replace_na(., "\U2014"))) %>%
  ## remove parentheses, but keep any content that was inside
  mutate(across(everything(), ~ str_replace(., "\\((.*)\\)", "\\1"))) %>%
  ## remove blank strings (^ = start, $ = end, so ^$ = start to end w/ nothing)
  mutate(across(everything(), ~ str_replace_na(., "^$"))) %>%
  ## fix years like above
  mutate(year = str_sub(year, 1, 4)) %>%
  ## convert to numbers, suppressing warnings about NAs b/c we know
  mutate(across(everything(), ~ suppressWarnings(as.numeric(.)))) %>%
  ## drop rows with missing year (blank online)
  drop_na(year)

## show
tab
```

Got it!

# Reshape data
```{r, echo = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Reshape data
## ---------------------------------------------------------

```

We could stop where we are, but to make the data more usable in
the future, let's convert to a long data frame. This takes a couple of
steps, but the idea is to have each row represent a year by estimate,
with a column for the estimate value and a column for the standard
error on that estimate. It may help to run the code below one line at
a time, checking the progress at each step.

```{r}
## gather for long data
df <- tab %>%
  ## pivot_longer estimates, leaving standard errors wide for the moment
  pivot_longer(cols = -c(year, ends_with("se")),
               names_to = "group",
               values_to = "estimate") %>%
  ## pivot_longer standard errors
  pivot_longer(cols = -c(year, group, estimate),
               names_to = "group_se",
               values_to = "se") %>% 
  ## drop "_se" from standard error estimates
  mutate(group_se = str_replace(group_se, "_se", "")) %>%
  ## filter where group == group_se
  filter(group == group_se) %>%
  ## drop extra column
  select(-group_se) %>%
  ## arrange
  arrange(year) %>%
  ## drop if missing year after reshaping
  drop_na(year)

## show
df
```

# Plot trends
```{r, echo = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Plot trends
## ---------------------------------------------------------

```

Let's look at overall college enrollment percentages for recent
graduates over time. Because our data are nicely formatted, it's easy
to subset the full table to data to only those estimates we need as
well as generate 95% confidence intervals.

```{r}
## adjust data for specific plot
plot_df <- df %>%
  filter(group %in% c("enr_pct", "enr_pct_m", "enr_pct_f")) %>%
  mutate(hi = estimate + se * qnorm(.975),
         lo = estimate - se * qnorm(.975),
         year = ymd(as.character(year), truncated = 2L),
         group = ifelse(group == "enr_pct_f", "Women",
                        ifelse(group == "enr_pct_m", "Men", "All")))

## show
plot_df
```

First, let's plot the overall average. Notice that we use the
`filter()` function in the `ggplot()` function to remove the subgroup
estimates for men and women. 

```{r scrape_trends_pct}
## plot overall average
g <- ggplot(plot_df %>% filter(group == "All"),
            mapping = aes(x = year, y = estimate)) +
  ## create shaded ribbon for 95% CI
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70") +
  ## line for main estimate
  geom_line() +
  ## make x-axis look nice
  ## major breaks: every 5 years, from min year to max year
  ## minor breaks: every 1 year, from min year to max year
  ## labels: formate to only show year ("%Y")
  scale_x_date(breaks = seq(min(plot_df$year),
                            max(plot_df$year),
                            "5 years"),
               minor_breaks = seq(min(plot_df$year),
                                  max(plot_df$year),
                                  "1 years"),
               date_labels = "%Y") +
  ## good labels and titles
  labs(x = "Year",
       y = "Percent",
       title = "Percent of recent high school completers in college: 1960 to 2016",
       caption = "Source: NCES Digest of Education Statistics, 2017, Table 302.10")    

## show
g
```

After a small dip in the early 1970s enrollment trends have steadily
risen over time.

Now let's compare enrollments over time between men and women
(dropping the overall average so our plot is clearer).

```{r scrape_trends_gender}
## plot comparison between men and women
g <- ggplot(plot_df %>% filter(group %in% c("Men","Women")),
            ## add colour == group to separate between men and women
            mapping = aes(x = year, y = estimate, colour = group)) +
  ## ribbon for 95% CI, but lower alpha so more transparent
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = group), alpha = 0.2) +
  ## primary estimate line
  geom_line() +
  ## neat x-axis breaks as before
  scale_x_date(breaks = seq(min(plot_df$year),
                            max(plot_df$year),
                            "5 years"),
               minor_breaks = seq(min(plot_df$year),
                                  max(plot_df$year),
                                  "1 years"),
               date_labels = "%Y") +
  ## good labels and titles
  labs(x = "Year",
       y = "Percent",
       title = "Percent of recent high school completers in college: 1960 to 2016",
       caption = "Source: NCES Digest of Education Statistics, 2017, Table 302.10") +
  ## set legend title, drop legend for colour since it's redundant with fill
  guides(fill = guide_legend(title = "Group"),
         colour = "none") +
  ## position legend so that it sits on plot face, in lower right-hand corner
  theme(legend.position = c(1,0), legend.justification = c(1,0))

## show
g
```

Though a greater proportion of men enrolled in college in the 1960s
and early 1970s, women have been increasing their enrollment
percentages faster than men since the 1980s and now have comparatively
higher rates of college participation.

#### Not-so-quick exercise
> Find the unemployment rate for 25 to 34 year-olds by degree type for
> the years 2014 through 2016. Make a long data frame and use it to
> make a figure of trends over time by educational
> attainment.  
> 
> See Table 501.10 of the NCES Digest of Education Statistics, which
> can can be found
> [here](https://nces.ed.gov/programs/digest/2017menu_tables.asp)).

```{r, echo = FALSE, purl = TRUE}

## =============================================================================
## END SCRIPT
################################################################################
```
