---
layout: lesson
title: "Data Visualization with ggplot2"
subtitle: "EDH7916 | Summer C 2020"
author: Benjamin Skinner
order: 6
category: lesson
links:
  script: plotting.R
  pdf: plotting.pdf
  data: 
  - hsls_small.dta
  - sch_test.zip
output:
  md_document:
    variant: gfm
    preserve_yaml: true
---

One key part of understanding your data and presenting your analyses
lies making plots. This lesson will focus on graphics.

R has three major graphing systems: the [base
system](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/00Index.html),
[lattice](http://lattice.r-forge.r-project.org), and
[ggplot2](http://ggplot2.tidyverse.org). Each system has its benefits
and drawbacks. Each is also very versatile with many, many options for
creating and adjusting plots.

Unfortunately, there isn't enough time to go through all three
graphing systems. After describing a few base R graphing functions,
this lesson will focus on using ggplot2 since it allows users to build
plots using the [grammar of
graphics](http://vita.had.co.nz/papers/layered-grammar.html) and
integrates well with the tidyverse.

# Setup

We're using two libraries today:

- [ggplot2](http://ggplot2.tidyverse.org)
- [haven](http://haven.tidyverse.org)

The [ggplot2](http://ggplot2.tidyverse.org) library is part of the
tidyverse, so we don't need to load it separately (we can
just use `library(tidyverse)` as always). 

We're also going to use [haven](http://haven.tidyverse.org), which
allows us to read in data files from other software such as SPSS, SAS,
and Stata. We'll use it to read in a Stata (`*.dta`) version of the
small HSLS data we've used before. The Stata version, unlike the plain
CSV version, has labels for the variables and values. These will be
useful when plotting. 

Though haven is part of the tidyverse (and should have been installed
when you installed tidyverse), we'll have to explicitly call it.

```{r, include = FALSE, purl = FALSE}
source('knit_setup.R')
```
```{r, include = FALSE, purl = TRUE}
################################################################################
##
## <PROJ> EDH7916: Data visualization with ggplot2
## <FILE> plotting.R 
## <INIT> 9 March 2020
## <AUTH> Benjamin Skinner (GitHub/Twitter: @btskinner)
##
################################################################################

```
```{r}
## ---------------------------
## libraries
## ---------------------------

library(tidyverse)
library(haven)
```

In addition to the Stata version of small HSLS, we'll also be using
`all_schools.csv` in the lesson. [As in the prior
lesson](../lessons/dw_two.html), unzip the file and place the entire
`sch_test` subdirectory will all included files in the `data`
subdirectory (if you don't already have it). After including its
subdirectory path (`tsc_dir`), we'll read in both data files.  

**Note** that since we have two data files this lesson, we'll give
them unique names instead of the normal `df`:

- `df_hs` := `hsls_small.dta`
- `df_ts` := `all_schools.csv`

```{r}
## ---------------------------
## directory paths
## ---------------------------

## assume we're running this script from the ./scripts subdirectory
dat_dir <- file.path("..", "data")
tsc_dir <- file.path(dat_dir, "sch_test")
```

```{r}
## ---------------------------
## input data
## ---------------------------

## assume we're running this script from the ./scripts subdirectory
## read_dta() ==> read in Stata (*.dta) files
## read_csv() ==> read in comma separated value (*.csv) files
df_hs <- read_dta(file.path(dat_dir, "hsls_small.dta"))
df_ts <- read_csv(file.path(tsc_dir, "all_schools.csv"))
```

# Plots using base R
```{r, echo = FALSE, purl = TRUE}
## -----------------------------------------------------------------------------
## Base R graphics
## -----------------------------------------------------------------------------
```

Even though new graphics libraries have been developed, the base R
graphics system remains powerful. The base system is also very easy to
use in a pinch. When I want a quick visual of a data distribution
that's just for me, I often use base R.

**Note** that for the next few plots, I'm not much concerned with how
they look. Specifically, the axis labels won't look very nice. We
could spend time learning to make really nice base R plots for
publication, but I'd rather we spend that time with ggplot2 graphics.

**Also note** that we'll switch to using the base R data frame `$`
notation to pull out the columns we want. If you need some more
information on using `$` notation, [check out the supplemental lesson
on data wrangling with base R](/edh7916/lessons/dw_one_base_r.html).

## Histogram
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## histogram
## ---------------------------
```

For continuous variables, a histogram is a useful plot. Though the
`hist()` function has many options to adjust how it looks, the
defaults work really well if you just want a quick look at the
distribution.

```{r eda_base_histogram}
## histogram of math scores (which should be normal by design)
hist(df_hs$x1txmtscor)
```

> #### Quick exercise
> Check the distribution of the students' socioeconomic score (SES).

## Density 
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## density
## ---------------------------
```

Density plots are also really helpful. R doesn't have single density
plot function, but you can get a density plot in one of two ways, each
of which will give a slightly different result.

First, you can adjust the `hist()` function to add the `freq = FALSE`
argument. It looks like the first histogram above, but notice that the
y-axis now represents density rather than counts.

```{r eda_base_hist_density}
## density plot of math scores with hist() function
hist(df_hs$x1txmtscor, freq = FALSE)
```

Second, you can `plot()` the `density()` of a continuous
variable. Unlike `hist()`, however, `density()` doesn't automatically
ignore missing values, so we have to tell it to remove `NA`s using the
`na.rm = TRUE` argument (a common argument for base R functions that's
useful to remember).

```{r eda_base_density}
## density plot of math scores
## read inside out: get density value, then plot values
plot(density(df_hs$x1txmtscor, na.rm = TRUE))
```

> #### Quick exercise
> Plot the density of SES. Next, add the `col`
> argument in `plot()` to change the color of the line to `"red"`:
> `plot(density(df_hs$x1txtmscor, na.rm = TRUE), col = "red")`.

## Box plot
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## box plot
## ---------------------------
```

A box plot will let you see the distribution of a continuous variable
at specific values of another discrete variable. For example, test
scores ranges at each student expectation level.

Call a box plot using the `boxplot()` function. This one is a little
trickier because it uses the [R
formula](https://www.statmethods.net/graphs/boxplot.html) construction
to set the continuous variable against the discrete variable. The formula uses a
tilde, `~`, and should be constructed like this:

* `<continuous var> ~ <discrete var>`

Notice how we can use the `data = df_hs` argument instead of adding
`df_hs$` in front of the variable names. This saves us some typing.

```{r eda_base_box}
## box plot of math scores against student expectations
boxplot(x1txmtscor ~ x1stuedexpct, data = df_hs)
```

From the boxplot, we can see that math test scores tend to increase as
students' educational expectations increase (remember that 11 means "I
don't know [how far I'll go in school]"), though there's quite a bit
of overlap in the marginal distributions.

## Scatter
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## scatter
## ---------------------------
```

Plot two continuous variables against one another using the base
`plot()` function. There are two primary ways to make a scatter plot
using `plot()`:

* `plot(x, y)`
* `plot(y ~ x)`

With both, `x` is the variable that will go on the x-axis and `y` the
one that will go on the y-axis. It's really a matter of which makes
sense to you. We'll use the first.

```{r eda_base_scatter}
## scatter plot of math against SES
plot(df_hs$x1ses, df_hs$x1txmtscor)
```

While the data seem to show a positive correlation between
socioeconomic status and math test score, there's also quite a bit of
variation in that association (notice that the cloud-like nature of
the circles). 

> #### Quick exercise 
> Rerun the above plot, but this time store it in an object,
> `plot_1`. Next, make the same plot, but this time use the second
> formula construction (`~`) --- store it in an object,
> `plot_2`. Visually compare the two, but for a more formal test, use
> `identical(plot_1, plot_2)` on the two plot objects to prove they
> are the same.

# Plots using ggplot2
```{r, echo = FALSE, purl = TRUE}
## -----------------------------------------------------------------------------
## Graphics with ggplot2
## -----------------------------------------------------------------------------
```

ggplot2 is my --- and many R users' --- primary system for making
plots. It is based on the idea of a [grammar of
graphics](https://www.springer.com/gp/book/9780387245447). Just as we
can use finite rules of a language grammar to construct an endless
number of unique sentences, so too can we use a few graphical
grammatical rules to make an endless number of unique figures.

The [ggplot2 system](http://ggplot2.tidyverse.org/reference/) is too
involved to cover in all of its details, but that's kind of the point
of the grammar of graphics: once you see how it's put together, you
can anticipate the commands you need to build your plot.

We'll start by covering the same plots as above.

## Histogram
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## histogram
## ---------------------------
```

[As the main help site says](http://ggplot2.tidyverse.org), all
ggplot2 plots need three things:

* **[data]**: The source of the variables you want to plot
* **[aesthetics]**: How variables in the data map onto the plot
  (_e.g._, what's on the x-axis? what's on the y-axis?)
* **[geom]**: The geometry of the figure or the kind of figure you
  want to make (_e.g._, what do you want to **do** with those data and
  mappings? A line graph? A box plot?...)

We'll start by making a histogram again. To help make these pieces
clearer, I'll use the argument names when possible. The first
function, which initializes the plot is `ggplot()`. Its first
argument is the data. 

The aesthetic mappings, that is, which variables go where or how they
function on the plot, go inside the `aes()` function. Since we only
have one variable, `x1txmtscor`, it is assigned to `x`.

If we assign this first part to an object, `p`, and print by calling
the object...

```{r eda_plot_histogram_blank, warning = F, message = F}
## init ggplot 
p <- ggplot(data = df_hs, mapping = aes(x = x1txmtscor))
p
```

...nothing! Well, not nothing, but no histogram. That's because the
plot object `p` knows the data and the key variable mapping but
doesn't know what do with them. What do we want?

Since we want a histogram, we add the `geom_histogram()` function to
the existing plot object with a plus sign(`+`). Once we do that, we'll
try to print the plot again...

```{r eda_plot_histogram, warning = F, message = F}
## add histogram instruction (notice we can add pieces using +)
p <- p + geom_histogram()
p
```

Success!

Let's repeat it the whole process, but without the middle step:

```{r eda_plot_histogram_2, warning = F, message = F}
## create histogram using ggplot
p <- ggplot(data = df_hs, mapping = aes(x = x1txmtscor)) +
    geom_histogram()
p
```

As you can see, the code to make a ggplot2 figure looks a lot like
what we've seen with other tidyverse libraries, _e.g._ dplyr. The key
difference between ggplot2 and dplyr, however, is that while dplyr
uses the pipe (`%>%`) to connect different functions, ggplot2 uses a
plus sign (`+`).

It may help you remember the difference:

- dplyr _moves_ output from left to the input in the right and so
  needs a **pipe** (`%>%`)
- ggplot2 _adds_ layer upon layer to build up the final figure and
  so needs a **plus sign** (`+`)

## Density
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## density
## ---------------------------
```

Unlike the base R graphics system, ggplot2 does have a density
plotting command, `geom_density()`. Instead of building up the figure
piecemeal, we'll go ahead and chain the geom to the first command and
print.

Notice how the function chain is the mostly the same as above, but (1)
written in a single linked chain and (2) using a different `geom_*()`
command at the end to indicate that we want something different.

```{r eda_plot_density, warning = F, message = F}
## density
p <- ggplot(data = df_hs, mapping = aes(x = x1txmtscor)) +
    geom_density()
p
```

> #### Quick exercise
> Make a density plot of SES.

If we want to superimpose the density plot over the histogram, we only
need chain the two commands together with a slight modification in how
the histogram is made. This way, the histogram and the density will be
on the same scale.

The change happens in the `geom_histogram()` function, where we add a
new mapping: `aes(y = ..density..)`. (**NOTE:** this is similar to what we
did above in base R to make a histogram on a density scale.)

```{r eda_plot_histdens, warning = F, message = F}
## histogram with density plot overlapping
p <- ggplot(data = df_hs, mapping = aes(x = x1txmtscor)) +
    geom_histogram(mapping = aes(y = ..density..)) +
    geom_density()
p
```

It worked, but it's not the greatest visual since the colors are the
same and the density plot is thin with no fill. 

Adding to what came before, the `geom_histogram()` and
`geom_density()` both take on new arguments that change the
defaults. Now the resulting plot should look nicer and be easier to
read.

```{r eda_plot_histdens_2, warning = F, message = F}
## histogram with density plot overlapping (add color to see better)
p <- ggplot(data = df_hs, mapping = aes(x = x1txmtscor)) +
    geom_histogram(mapping = aes(y = ..density..),
                   color = "black",
                   fill = "white") +
    geom_density(fill = "red", alpha = 0.2)
p
```

> #### Quick exercise
> Try changing some of the arguments in the last plot. What happens
> when you change `alpha` (keep the value between 0 and 1)? What does
> the `color` argument change? And `fill`? What happens if you switch
> the `geom_*()` functions, call `geom_histogram()` after you call
> `geom_density()`?

**A key thing to note about arguments** is that when the are outside
of the `aes()`, they apply uniformly to the whole geom (_e.g._ all the
histogram bars are white with a black outline, the density is light
red). When you want some aesthetic of the figure to change as a
function of the data, you need to put it inside `aes()`. We'll see
this in the next plot.

## Two-way
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## two way plot
## ---------------------------
```

Plotting the difference in a continuous distribution across groups is
a common task. Let's see the difference between student math scores
between students with parents who have any postsecondary degree and
those without. 

Since we're using data that was labeled in Stata, we'll see the
labels when we use `count()`

```{r}
## see the counts for each group
df_hs %>% count(x1paredu)
```

We can see that all values of `x1paredu` greater than 2 represent
parents with some college credential. Since we want only two distinct
groups, we can use the operator `>=` to make a new 0/1 binary
variable. If a value of `x1paredu` is above 3, then the new indicator
`pared_coll` will be 1; if not, 0.

**NOTE** that in the Stata version of `hsls_small`, all the missing
values, which are normally negative numbers, have already been
properly converted to `NA` values. That's why we see a count column
for `NA` and not labels for missingness that we might have expected
based on prior lessons.

The `ggplot()` function doesn't need to use our full data. In fact,
our data needs to be set up a bit differently to make this plot. We'll
make a new temporary data object that only has the data we need.


```{r}
## need to set up data
plot_df <- df_hs %>%
    ## select the columns we need
    select(x1paredu, x1txmtscor) %>%
    ## can't plot NA so will drop
    drop_na() %>%
    ## create new variable that == 1 if parents have any college
    mutate(pared_coll = ifelse(x1paredu >= 3, 1, 0)) %>%
    ## drop (using negative sign) the original variable we don't need now
    select(-x1paredu) 

## show
head(plot_df)
```

To plot against the two groups we've made, we need to add it to the
aesthetic feature, `aes()`. The math score, `x1txmtscor`, is still
mapped to `x`, but since we want two side-by-side histograms, we set
the `fill` aesthetic to our new indicator variable. So the function
knows that it's a group (and not just a continuous number with only
two values), we wrap it in the `factor()` function.

Finally, we add some changes to the `geom_histogram()` function so that
each group is on the same scale. 

```{r eda_plot_histogram_double, warning = F, message = F}
## two way histogram
p <- ggplot(data = plot_df,
            aes(x = x1txmtscor, fill = factor(pared_coll))) +
    geom_histogram(alpha = 0.5, stat = "density", position = "identity")
p
```

By assigning `pared_coll` to the `fill` aesthetic, we can see a
difference in the distribution of math test scores between students
whose parents have at least some college and those whose parents do
not. 

> #### Quick exercise
> Remove some of the new arguments in `geom_histogram()`. How does the
> resulting plot change? Remove the `factor()` function from around
> `pared_coll`: what happens?

## Box plot
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## box plot
## ---------------------------
```

By this point, you're hopefully seeing the pattern in how ggplot2
figures are put together. To make a box plot, we need to add a `y`
mapping to the `aes()` in addition to the `x` mapping. We've also
added the same variable to `fill` as we did to `x`.  We do this so
that in addition to having different box and whisker plots along the
x-axis, each plot is given its own color.

```{r eda_plot_box, warning = F, message = F}
## box plot using both factor() and as_factor()
p <- ggplot(data = df_hs,
            mapping = aes(x = factor(x1paredu),
                          y = x1txmtscor,
                          fill = as_factor(x1paredu))) +
    geom_boxplot()
p
```

In a way, this plot is similar to the dual histogram above. But since
we want to see the distribution of math scores across finer-grained
levels of parental education, the box and whisker plot is clearer than
trying to overlap seven histograms.

> #### Quick exercise
> Change the `as_factor()` and `factor()` functions above. How does
> the plot change?

# Scatter
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## scatter plot
## ---------------------------
```

To make a scatter plot, make sure that the `aes()` has mappings for
the `x` axis and `y` axis and then use `geom_point()` to plot. To make
things easier to see (remembering the cloud from the base R plot
above), we'll reduce the data to 10% of the full sample using
`sample_frac()` from dplyr. We'll also limit our 10% to those who
aren't missing information about student education expectations

```{r eda_plot_scatter_1, warning = F, message = F}
## sample 10% to make figure clearer
df_hs_10 <- df_hs %>%
    ## drop observations with missing values for x1stuedexpct
    drop_na(x1stuedexpct) %>%
    ## sample
    sample_frac(0.1)

## scatter
p <- ggplot(data = df_hs_10, mapping = aes(x = x1ses, y = x1txmtscor)) +
    geom_point()
p

```

Now that we have our scatter plot, let's say that we want to add a
third dimension. Specifically, we want to change the color of each
point based on whether a student plans to earn a Bachelor's degree or
higher. That means we need a new dummy variable that is 1 for those
with BA/BS plans and 0 for others.

We can look at the student base year expectations with `count()`:

```{r}
## see student base year plans
df_hs %>%
    count(x1stuedexpct)
```

We see that `x1stuedexpct >= 6` means a student plans to earn a
Bachelor's degree or higher. But since we need to account for the fact
that 11 means "I don't know", we need to make sure our test includes
`x1stuedexpct < 11`. Remember from a prior lesson that we can connect
these two statements together with the operator `&`. Let's create our
new variable.

```{r}
## create variable for students who plan to graduate from college
df_hs_10 <- df_hs_10 %>%
    mutate(plan_col_grad = ifelse(x1stuedexpct >= 6 & x1stuedexpct < 11,
                                  1,        # if T: 1
                                  0))       # if F: 0
```

Now that we have our new variable `plan_col_grad`, we can add it the
`color` aesthetic, `aes()` in `geom_point()`. Don't forget to use
`factor()` so that ggplot knows to treat it like a group!

```{r eda_plot_scatter_2, warning = F, message = F}
## scatter
p <- ggplot(data = df_hs_10,
            mapping = aes(x = x1ses, y = x1txmtscor)) +
    geom_point(mapping = aes(color = factor(plan_col_grad)), alpha = 0.5)
p
```

> #### Quick exercise
> Change how you make `plan_col_grad` so that instead of 1 and 0, you
> use 'yes' and 'no'. Make your figure again. What changes?

## Fitted lines
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## fitted lines
## ---------------------------
```

It's often helpful to plot fitted lines against a scatter plot to help
see the underlying trend. There are a number of ways to do this with
the `geom_smooth()` function.

### Linear fit

Setting `method = lm` in `geom_smooth()` will fit a simple straight
line of best fit with 95% confidence interval shaded around it.

```{r eda_plot_scatter_3, warning = F, message = F}
## add fitted line with linear fit
p <- ggplot(data = df_hs_10, mapping = aes(x = x1ses, y = x1txmtscor)) +
    geom_point(mapping = aes(color = factor(plan_col_grad)), alpha = 0.5) +
    geom_smooth(method = lm)
p
```

### Linear fit with polynomials

In addition to the `method`, we can add a `formula` to allow the
fitted line to take a non-linear shape. Using the `aes()` values of
`x` and `y`, the argument below uses an R formula, `y ~ x`, but with
the addition of the `poly()` function. Setting the second argument in
`poly()` to 2 gives the line an extra quadratic term, which allows it
to take a more curved shape.

```{r eda_plot_scatter_4, warning = F, message = F}
## add fitted line with polynomial linear fit
p <- ggplot(data = df_hs_10, mapping = aes(x = x1ses, y = x1txmtscor)) +
    geom_point(mapping = aes(color = factor(plan_col_grad)), alpha = 0.5) +
    geom_smooth(method = lm, formula = y ~ poly(x,2))
p
```

> #### Quick exercise
> Change the value in `poly()` to higher numbers. How does the line
> change? 

### Loess

Finally, we can skip trying to adjust a linear line and just fit a
[LOESS](https://en.wikipedia.org/wiki/Local_regression) curve, which
is a smooth line produced by fitting a large number of local
polynomial regressions on subsets of the data.

```{r eda_plot_scatter_5, warning = F, message = F}
## add fitted line with loess
p <- ggplot(data = df_hs_10, mapping = aes(x = x1ses, y = x1txmtscor)) +
    geom_point(mapping = aes(color = factor(plan_col_grad)), alpha = 0.5) +
    geom_smooth(method = loess)
p
```

To be clear, these semi-automated lines of best fit should not be used
to draw final conclusions about the relationships in your data. You
will want to do **much more** analytic work to make sure any
correlations you observe aren't simply spurious and that fitted lines
are telling you something useful. That said, fitted lines via
ggplot2 can be useful when first trying to understand your data or
to more clearly show observed trends.

## Line graph
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## line graph
## ---------------------------
```

When you want to show changes in one variable as a function of another
variable, _e.g._, changes in test scores over time, then a line graph
is typically your best choice. Since our `hsls_small` data is
cross-sectional, we'll shift to using our school test score
data. Remember that the test score data show three sets of test scores
(math, science, and reading) for four schools over a period of six
years. This data frame is long in year, but wide in test type. It
looks like this:

```{r}
## show test score data
df_ts
```

To keep it simple for our first line plot, we'll filter our plot data
to keep only scores for one school. Notice how we can do that directly
with pipes inside the `ggplot()` function. We want to see changes in
test scores over time, so we'll map `year` to the `x` axis and, for
now, `math` to the `y` axis. To see a line graph, we add
`geom_line()`.

```{r, warning = F, message = F}
## line graph
p <- ggplot(data = df_ts %>% filter(school == "Spottsville"),
            mapping = aes(x = year, y = math)) +
    geom_line()
p
```

> #### QUICK EXERCISE 
> Change the school in `filter()` to "East Heights" and then "Bend
> Gate".

Easy enough, but let's say that we want to add a third dimension ---
to show math scores for each school in the same plot area. To do this,
we can map a third aesthetic to `school`. [Looking at the help file
for
`geom_line()`](https://ggplot2.tidyverse.org/reference/geom_path.html#aesthetics),
we see that lines (a version of a path) can take `colour`, which means
we can change line color based on a variable. 

The code below is almost exactly the same as before less two things:

1. We don't filter `df_ts` this time, because we want all schools
2. We add `colour = school` inside `aes()`

```{r, warning = F, message = F}
## line graph for math scores at every school over time
p <- ggplot(data = df_ts,
            mapping = aes(x = year, y = math, colour = school)) +
    geom_line()
p
```

This is nice (though maybe a little messy at the moment) because it
allows us to compare math scores across time across schools. But we
have two more test types --- reading and science --- that we would
like to include as well. One approach that will let us add yet another
dimension is to use facets.

## Facets

With facets, we can put multiple plots together, each showing some
subset of the data. For example, instead of plotting changes in math
scores across schools over time on the same plot area (only changing
the color), we can use `facet_wrap()` to give each school its own
little plot. You might hear me or other refer to plots like this a
showing _small multiples_ or as _small multiples_ figures.

Compared to the code just above, notice how we've removed
`colour = school` from `aes()` and included `facet_wrap(~
school)`. The tilde (`~`) works like the tilde in `plot(y ~ x)` above:
it means "plot against or by _X_". In this case, we are plotting math
test scores over time _by_ each school.

```{r, warning = F, message = F}
## facet line graph
p <- ggplot(data = df_ts,
            mapping = aes(x = year, y = math)) +
    facet_wrap(~ school) +
    geom_line()
p
```

Is this faceted plot better than the color line plot before it? To my
eyes, it's a little clearer, but not so much so that I couldn't be
convinced to use the first one. Whether you use the first or the
second would largely depend on your specific data and presentation
needs.

Faceting has a clearer advantage, however, when you want to include
the fourth level of comparison: (1) scores across (2) time across (3)
schools from (4) different tests. To make this comparison, we first
need to reshape our data, which is only long in `year`, to be long in
`test`, too. As we've already seen in a past lesson, we'll use
`pivot_longer()` to place each test type in its own column (`test`)
with the `score` next to it.


```{r}
## reshape data long
df_ts_long <- df_ts %>%
    pivot_longer(cols = c("math","read","science"), # cols to pivot long
                 names_to = "test",                 # where col names go
                 values_to = "score")               # where col values go

## show
df_ts_long
```

> #### QUICK EXERCISE
> If we have 4 schools, 6 years, and 3 tests, how many observations
> should `df_ts_long` have in total? Does it?

With our reshaped data frame, we now reintroduce `colour` into the
`aes()`, this time set to `test`. We make one other change: `y =
score` now, since that's the column for test scores in our reshaped
data. All else is the same.

```{r, warning = F, message = F}
## facet line graph, with colour = test and ~school
p <- ggplot(data = df_ts_long,
            mapping = aes(x = year, y = score, colour = test)) +
    facet_wrap(~ school) +
    geom_line()
p
```

Well, it worked...we can see each school's different test score trends
over time, with each school in its own facet and test scores set to a
different color. But the result is a bit underwhelming. Because the
different test types are such different scales (even though they are
normed within themselves), within-test changes seem rather flat over
time.

Let's try something different: in the next figure, we'll swap the
variables we give to `colour` and within `facet_wrap()`. This means
that each test should have its own facet and each line will represent
a different school.

```{r, warning = F, message = F}
## facet line graph, now with colour = school and ~test
p <- ggplot(data = df_ts_long,
            mapping = aes(x = year, y = score, colour = school)) +
    facet_wrap(~ test) +
    geom_line()
p
```

Okay. New problem. While it's maybe a _little_ easier to see same-test
differences across schools over time, the different scales of the
tests still make the figure less useful than we might hope. It's not
that the students are _way_ better at science than reading; it's just
that the tests are scaled differently. Someone quickly reading this
figure, however, might make that incorrect interpretation.

One thing we can do is change the y-axis for each facet. The default is
to keep the y-axis scale the same. By adding `scales = "free_y"` to
`facet_wrap()`, we'll let each test have its own y-axis scale.

Having different axis scales side-by-side can be confusing, however
(this is why the default is to keep them the same). To mitigate that
confusion, we'll also rearrange the facets so they stack rather than
sit side by side. To do this, we'll add `ncol = 1` to
`facet_wrap()`. This says our facets have to stick to one column,
effectively meaning they will stack vertically.

```{r, warning = F, message = F}
## facet line graph, with one column so they stack
p <- ggplot(data = df_ts_long,
            mapping = aes(x = year, y = score, colour = school)) +
    facet_wrap(~ test, ncol = 1, scales = "free_y") +
    geom_line()
p
```

That looks better! But we can do even better than that...

Currently, each test score is on its own normed scale. While our new
figure allows us to make comparisons across schools over time _within_
test, it's more difficult to make a good comparison _between_
tests. For example, East Heights has a little over 20 point drop in
reading scores from 1981 to 1982 and about the same drop in science
scores from 1982 to 1983. How should we think about these drops? Are
they about the same or is one drop relatively bigger than the other?

To better answer this question, we could re-standardize each test
score so that it is centered at 0 and a one unit change is equal to 1
standard deviation difference in score. We'll use `mutate()` to create
a new variable `score_std`. Because we `group_by()` `test`,
`score_std` will be standardized within test.

```{r}
## rescale test scores
df_ts_long <- df_ts_long %>%
    group_by(test) %>%
    mutate(score_std = (score - mean(score)) / sd(score)) %>%
    ungroup
```

We'll repeat the same code as before, but this time substitute `y =
score_std`. Because all tests are on the same standardized scale, we
can also drop `scales = "free_y"`.

```{r, warning = F, message = F}
## facet line graph with standardized test scores
p <- ggplot(data = df_ts_long,
            mapping = aes(x = year, y = score_std, colour = school)) +
    facet_wrap(~ test, ncol = 1) +
    geom_line()
p
```

Notice the lines look the same relative to one another, but now we
have a consistent scale to help judge changes. To answer our question
from before, it seems that the drop in reading scores (1981 to 1982)
and science scores (1982 to 1983) were each about 1.5 standard
deviations. We could test more formally, but we have a clearer idea
now that all tests are on the same scale.

> #### QUICK EXERCISE
> What happens if you use the argument `scales = "free_y"` in the last
> bit of code? Why might you not use that once we've scaled the test
> scores?

As a quick change, we can go back to having each school in its own
facet and test scores within.

```{r, warning = F, message = F}
## facet line graph
p <- ggplot(data = df_ts_long,
            mapping = aes(x = year, y = score_std, colour = test)) +
    facet_wrap(~ school) +
    geom_line()
p
```

> #### QUICK EXERCISE
> Why did we drop `ncol = 1` from `facet_wrap()`? What happens if you
> keep it?

Our plot is looking better, but it still may not contain the
information we want. We've standardized the test scores over this time
window, but maybe what we really want to know is how they've _changed
relative to the beginning of the sample period_. You can imagine a
superintendent who took over in 1981 would be keen to know how scores
have changed during their tenure.

This means that while we still want to standardize the scores, we
should zero them not at the overall mean, but at the value in the first
year. We can do that by grouping by `school` and `test`, arranging in
year order, making a new variable that is the `first()` score (within
test, within school) and using that rather than the mean test score to
make our new variable, `score_std_sch`.

```{r}
## standardize test scores within school to first year
df_ts_long <- df_ts_long %>%
    group_by(test, school) %>%
    arrange(year) %>% 
    mutate(score_year_one = first(score),
           ## note that we're using score_year_one instead of mean(score)
           score_std_sch = (score - score_year_one) / sd(score)) %>%
    ungroup
```

Now we'll plot using our new variable `score_std_sch`.

```{r, warning = F, message = F}
## facet line graph
p <- ggplot(data = df_ts_long,
            mapping = aes(x = year, y = score_std_sch, colour = test)) +
    facet_wrap(~ school) +
    geom_line()
p
```

With this final graph, we can see relative changes across schools,
across times, across tests. Notice that line shapes within each facet
are the same as before, just shifted up or down so that the first
value for each test in 1981 is 0. 

Is this the best version of this figure (minus making the axis and
legend labels look nicer)?  Again, it depends on what you want to
show. Remember that figures don't speak for themselves: it's up to you
to explain to your reader (include your future self) what they
mean. That said, a well crafted figure will make that job much easier.

```{r, echo = FALSE, purl = TRUE}

## =============================================================================
## END SCRIPT
################################################################################
```
