```{r setup, echo=F, include=F, message=F, warning=F, error=F}
## ---------------------------
## libraries
## ---------------------------

library(knitr)
library(tidyverse)

## ---------------------------
## settings/macros
## ---------------------------

## NB:
## - echo (FALSE): don't repeat this code in output
## - include (FALSE): run code, but don't include output (unless a plot)
## - message (FALSE): don't output any messages
## - warning (FALSE): don't output any warnings
## - error (FALSE: don't output any errors
##
## We'll include these in the general knitr::opts_chunk() below, but
## need them here to silence unnecessary output before we can set the options

## set up knitr options
opts_chunk$set(error = FALSE,
               echo = FALSE,
               include = FALSE,
               message = FALSE,
               warning = FALSE,
               fig.path = "../figures/repo-", # where figures should be stored
               dpi = 300,                  # print quality (300 standard for print)
               out.width = "100%",         # figures should be as wide as margins
               comment = NA)               # if code output, no comment char on LHS

## ---------------------------
## directory paths
## ---------------------------

## read in our data here, assuming we're in scripts like always
dat_dir <- file.path("..", "data")
sch_dir <- file.path(dat_dir, "sch_test")
```

```{r input}
## ---------------------------
## input
## ---------------------------

df <- read_csv(file.path(sch_dir, "all_schools.csv"))

```

From 1980 to 1985, students at four schools took end of year exams in
three subjects --- math, reading, and science. While these tests did
not affect students' grades or promotion, they were meant to measure
what students had learned over the course of the school year. In each
year, only 9th grade students took the exam. This means that each year
of data represents a different cohort of 9th grade students. Because
test scores are standardized within subject area, student cohorts can
be compared across time. The table below shows average test scores for
each school in each year.

```{r table_all, include = T}
## ---------------------------
## make table of all scores
## ---------------------------

## use the kable() function in knitr to make nicer table
kable(df,
      digits = 0,
      col.names = c("", "Year", "Math", "Reading", "Science"))
```

The tests are scaled as follows (observed averages and standard
deviations are somewhat different due to natural variation):

- **Math** --- mean: 500; standard deviation: 10
- **Reading** --- mean: 300; standard deviation: 20
- **Science** --- mean: 800; standard deviation: 15

# Test score averages by school

```{r table_averages, include = T}
## ---------------------------
## make table of averages
## ---------------------------

df_tab <- df %>%
    ## group by school
    group_by(school) %>%
    ## get average across years
    summarise(math_mean = mean(math),
              read_mean = mean(read),
              science_mean = mean(science))

## store variables to use in text below
hi_math_sch <- df_tab %>% filter(math_mean == max(math_mean)) %>% pull(school)
hi_math_scr <- df_tab %>% filter(math_mean == max(math_mean)) %>% pull(math_mean)

hi_read_sch <- df_tab %>% filter(read_mean == max(read_mean)) %>% pull(school)
hi_read_scr <- df_tab %>% filter(read_mean == max(read_mean)) %>% pull(read_mean)

hi_sci_sch <- df_tab %>% filter(science_mean == max(science_mean)) %>% pull(school)
hi_sci_scr <- df_tab %>% filter(science_mean == max(science_mean)) %>% pull(science_mean)

## use the kable() function in knitr to make nicer table
kable(df_tab,
      digits = 0,
      col.names = c("", "Math", "Reading", "Science"))
```

Across the six years of data, `r hi_math_sch` had the
highest average math score (`r hi_math_scr %>% round`); 
`r hi_read_sch` had the highest average reading score 
(`r hi_read_scr %>% round`); and 
`r hi_sci_sch` had the highest average science score 
(`r hi_sci_scr %>% round`). However, these six year averages cover a fair
amount of variation within schools across time. In the next sections, I'll
investigate this variation.

```{r reshape}
## ---------------------------
## reshape for figure
## ---------------------------

## reshape data long
df_long <- df %>%
    pivot_longer(cols = c("math","read","science"), # cols to pivot long
                 names_to = "test",                 # where col names go
                 values_to = "score")               # where col values go

```

# Test score trends

```{r fig_unadjusted, include = T}
## ---------------------------
## fig: unadjusted
## ---------------------------

## facet line graph, with one column so they stack
p <- ggplot(data = df_long,
            mapping = aes(x = year, y = score, colour = school)) +
    facet_wrap(~ test, ncol = 1, scales = "free_y") +
    geom_line()
p

```

Considering test score trends over time, shown in the figure above, I
note three important points:

1. First, test scores varied across schools in each testing year. This
is true of all tests.
2. Second, test score varied within schools across time. Again, this
   is true across tests.
3. Third, no single school consistently performed better or worse than
   other schools. Each school had up and down periods and when
   performing well on one test, often performed less well on another.
   

# Test score trends: adjusted to 1980

```{r add_first}
## ---------------------------
## standardize to first year
## ---------------------------

df_long <- df_long %>%
    group_by(test, school) %>%
    arrange(year) %>% 
    mutate(score_year_one = first(score),
           score_std_sch = (score - score_year_one) / sd(score)) %>%
    ungroup

```
```{r fig_adjusted, include = T}
## ---------------------------
## fig: adjusted to first year
## ---------------------------

## facet line graph
p <- ggplot(data = df_long,
            mapping = aes(x = year, y = score_std_sch, colour = test)) +
    facet_wrap(~ school) +
    geom_line()
p

```

All tests were revamped for the year 1980 to account for a significant
curriculum change in the 1979-1980 school year. This is why data begin
in this year, even though the district tested 9th graders in math,
reading, and science from 1970. Because of these changes, it makes
sense to compare score trends from a starting point of 1980.

Two schools, Bend Gate and Niagara, show a general downward trend in
math scores over time. East Heights, after an initial drop in math
scores from 1980 to 1981, however, steadily increased students' math
scores over time. East Heights also had generally strong science
scores, though with a dip after an initial increase. Niagara should
also be noted for its increasing science scores after an initial
drop. Though there is a large single year dip in math scores for
Spottsville in 1984, I note that the drop may be attributed to a
school emergency that interrupted testing on the day the math exam was
given.

# Recommendations

Despite variation across years, there is evidence that each school
tended to peform better (or worse) in some subject areas than in
others. Furthermore, schools differed in their relative strengths (and
weaknesses). Though more formal analyses are warranted, these
descriptive analyses suggest that district students may benefit from
more cross-school interaction among teachers, particularly so that
successful subject-specific practices and lessons may be replicated
in classrooms across the district.
